---
title: "Lima Bean GxE"
format: html
editor: visual
---

```{r}
library(reshape2)
library(ggplot2)
library(openxlsx)
library(asreml)
library(dplyr)
library(tibble)
library(ggplot2)
library(metan)
library(writexl)
```

## Data

-   3 Locations (Brazil) - Piracicaba - SP, Teresina - PI, Tianguá - CE,

-   40 Lima Bean Lines

-   Randomized Complete Block Design (RCBD) - 3 Repetitions

-   Plot usable area: 5m²

-   Traits:

    |          Trait           | Acronym | Unit  |
    |:------------------------:|:-------:|:-----:|
    |       Grain Yield        |   GY    | kg/ha |
    | Number days to flowering |   NDF   | days  |
    | Number days to maturity  |   NDM   | days  |
    | One Hundred Seed Weight  |  OHSW   |   g   |
    |       Plant Height       |   PH    |  cm   |
    |        Pod Length        |   PL    |  mm   |
    |        Pod Number        |   PN    |  mm   |
    |      Pod Thickness       |   PT    |  mm   |
    |        Pod Width         |   PW    |  mm   |
    |       Seed Length        |   SL    |  mm   |
    |      Seed Thickness      |   ST    |  mm   |
    |        Seed Width        |   SW    |  mm   |

```{r}
LB = read.xlsx("C://Users//joaop//Desktop//fava//LimaBean.xlsx")
LB$Block    <- as.factor(LB$Block)
LB$Env    <- as.factor(LB$Env)
LB$Genotype <- as.factor(LB$Genotype)
LB$GEN<- as.factor(LB$GEN)
str(LB)
```

## anova

```{r}
aov = anova_joint(LB,env = Env, gen = GEN, rep = Block,resp = everything())
```

## Mixed-effect models

```{r}
mixedmodel <- gamem_met(LB,env = Env, gen = GEN, rep = Block,random = "all",resp = everything())
```

#### LRT

```{r}
GYLRT = mixedmodel$GY$LRT
OHSWLRT = mixedmodel$OHSW$LRT 
NDFLRT = mixedmodel$NDF$LRT
NDMLRT = mixedmodel$NDM$LRT 
PHLRT = mixedmodel$PH$LRT
PNLRT = mixedmodel$PN$LRT
LRT = list(
  GY_LRT   = GYLRT,
  OHSW_LRT = OHSWLRT,
  NDF_LRT  = NDFLRT,
  NDM_LRT  = NDMLRT,
  PH_LRT   = PHLRT,
  PN_LRT   = PNLRT
)
LRT
write_xlsx(LRT, "LRT_LB.xlsx")
```

#### BLUPs

```{r}
BLUPgen = list(
  GY_BLUP   = mixedmodel$GY$BLUPgen,
  OHSW_BLUP = mixedmodel$OHSW$BLUPgen,
  NDF_BLUP  = mixedmodel$NDF$BLUPgen,
  NDM_BLUP  = mixedmodel$NDM$BLUPgen,
  PH_BLUP   = mixedmodel$PH$BLUPgen,
  PN_BLUP   = mixedmodel$PN$BLUPgen
)
write_xlsx(BLUPgen, "BLUPgen_LB.xlsx")
```

```{r}
library(ggplot2)
library(dplyr)

plot_trait_bar <- function(x, trait_name, size.tex.lab = 12) {
  
  df <- x[[trait_name]]$BLUPgen %>%
    mutate(
      Status = ifelse(Predicted < mean(Predicted, na.rm = TRUE), "Below", "Above")
    )
  
  df$Status <- factor(df$Status, levels = c("Below", "Above"))
  
  df <- df %>%
    arrange(Predicted) %>%
    mutate(GEN = factor(GEN, levels = GEN))
  
  ggplot(df, aes(x = GEN, y = Predicted, fill = Status)) +
    geom_col(color = "white", alpha = 0.4) +
    scale_fill_manual(values = c("Above" = "purple","Below" = "orange")) +
    coord_flip() +
    labs(
      x = "Ranking Genotypes",
      y = paste(trait_name)
    ) +
    theme_classic(base_size = 12) +
    theme(
      axis.text.y = element_text(size = 12),
      legend.position = "right"
    )
}
traits <- c("GY", "OHSW", "PH", "PN", "NDF", "NDM")
BLUP_bar_plots <- lapply(traits, function(trait) plot_trait_bar(mixedmodel, trait))
```

```{r}
library(ggpubr)
blups_panel3 = ggarrange(BLUP_bar_plots[[1]], BLUP_bar_plots[[2]], BLUP_bar_plots[[3]], BLUP_bar_plots[[4]], BLUP_bar_plots[[5]], BLUP_bar_plots[[6]], ncol = 3, nrow = 2, common.legend = T, legend = "bottom", labels = c("A", "B", "C", "D", "E", "F"), font.label = list(size = 10, face = "bold"))
blups_panel3
ggsave("BLUPs_LB.pdf",blups_panel3,width = 12, height = 14, dpi = 600)
```

#### Correlation

```{r}
corr = corr_coef(LB, NDF, NDM, PH, PN, GY, OHSW)
corrplot = plot(corr) +
  scale_fill_gradientn(
    colours = c("orange", "white", "purple"),
    limits = c(-1, 1)
  )
corrplot
ggsave("Correlation.pdf",corrplot, width = 4, height = 4, dpi = 600)
```

#### Genotype by trait biplot

```{r}
gtb1 = gtb(LB, GEN, resp = c(NDF, NDM, PH, PL, PW, 
             PN, GY, OHSW))

gtb1plot = plot(gtb1,
     col.env = "orange",
            col.gen = "purple",
            size.text.gen = 5,
            size.text.env = 8, repel = T,
            repulsion = 1,
            col.line = "orange",
            plot_theme = theme_metan_minimal())
ggsave("Genotypebytraitbiplot_LB.pdf",gtb1plot, width = 8, height = 8, dpi = 600)
```

##### Correlation + GYT

```{r}
lbplotcorrgyt = ggarrange(corrplot,gtb1plot, ncol = 2, nrow = 1, labels = c("A","B"), font.label = list(size = 10, face = "bold"))
lbplotcorrgyt
ggsave("Correlation_GYT_LB.pdf",lbplotcorrgyt,width = 12, height = 6, dpi = 600)
```

#### Which-won-Where

```{r}
dir.create("Which_won_where", showWarnings = FALSE)

# Loop por trait
for(tr in traits){
  
  # Calcular GGE para o trait
  gge3_tr <- gge(
    LB,
    Env,
    GEN,
    resp = tr,
    svp = "symmetrical"
  )
  
  # Plot type 3 (Which-won-where)
  p <- plot(gge3_tr,
            type = 3,
            col.env = "orange",
            col.gen = "purple",
            arrow.color = "purple",
            size.text.gen = 2,
            size.text.env = 5,
            size.shape.win = 4,
            large_label = 6,
            repel = TRUE,
            size.text.win = 4.5,
            repulsion = 2,
            col.line = "orange",
            plot_theme = theme_metan_minimal())
    ggsave(
    filename = paste0("Which_won_where/", tr, "_WhichWonWhere.pdf"),
    plot = p,
    width = 10,
    height = 8,
    dpi = 600
    )}
p
```

```{r}
# Plot type 3 (Which-won-where)
www <- list()

for(tr in traits){
  gge3_tr <- gge(LB, Env, GEN, resp = tr, svp = "symmetrical")
  
  p <- plot(gge3_tr, type = 3,
            col.env = "orange",
            col.gen = "purple",
            arrow.color = "purple",
            size.text.gen = 1,
            size.text.env = 5,
            size.shape.win = 5,
            large_label = 1,
            repel = T,
            size.text.win = 5,
            repulsion = 0.5,
            col.line = "orange",
            plot_theme = theme_metan_minimal())
  
  www[[tr]] <- p + ggtitle(NULL) + labs(subtitle = NULL)}

www_lb = ggarrange(www[["GY"]], www[["OHSW"]], www[["PN"]],
                   www[["PH"]], www[["NDF"]], www[["NDM"]],
                   ncol = 3, nrow = 2,
                   common.legend = T, legend = "bottom", 
                   labels = c("A", "B", "C", "D", "E", "F"), 
                   font.label = list(size = 10, face = "bold"))
www_lb <- annotate_figure(
  www_lb,
  top = text_grob(
    "Which-Won-Where",
    face = "bold",
    size = 16
  )
)
www_lb
```

```{r}
library(ggpubr)
library(ggplot2)

www <- list()

for(tr in traits){
  gge3_tr <- gge(LB, Env, GEN, resp = tr, svp = "symmetrical")
  
  p <- plot(gge3_tr, type = 3,
            col.env = "orange",
            col.gen = "purple",
            arrow.color = "purple",
            size.text.gen = 1,
            size.text.env = 4,
            size.shape.win = 4,
            large_label = 1,
            repel = TRUE,
            size.text.win = 4,
            repulsion = 0.5,
            col.line = "orange",
            plot_theme = theme_metan_minimal())
  
  p_try <- try({
    p2 <- p + labs(title = tr, subtitle = NULL) +
            theme(plot.title = element_text(hjust = 0.5, face = "bold"))
    p <- p2
    NULL
  }, silent = TRUE)
  
  if(inherits(p_try, "try-error") || !inherits(p, "ggplot")) {
    p_grob <- ggpubr::as_ggplot(p)            
    p <- p_grob + labs(title = tr, subtitle = NULL) +
              theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  }
  
  www[[tr]] <- p
}

www_lb <- ggarrange(
  www[["GY"]],  www[["OHSW"]], www[["PN"]],
  www[["PH"]],  www[["NDF"]],  www[["NDM"]],
  ncol = 3, nrow = 2,
  common.legend = TRUE, legend = "bottom",
  labels = c("A","B","C","D","E","F"),
  font.label = list(size = 10, face = "bold")
)

www_lb <- annotate_figure(
  www_lb,
  top = text_grob("Which-Won-Where", face = "bold", size = 16)
)

print(www_lb)
ggsave("Which-won-Where.pdf", www_lb, width = 10, height = 8, dpi = 600)
```

#### Mean_performance vs. stability 

```{r}
dir.create("Mean_performance_vs_stability", showWarnings = FALSE)

mps_list <- list()

for(tr in traits){
    gge2_tr <- gge(
    LB,
    Env,
    GEN,
    resp = tr,
    svp = "genotype"
  )
  
  # Plot type 2
  p <- plot(gge2_tr,
            type = 2,
            col.env = "orange",
            col.gen = "purple",
            size.text.gen = 2,
            size.text.env = 4,
            repel = TRUE,
            repulsion = 2,
            col.line = "orange",
            plot_theme = theme_bw())
  
  p_try <- try({
    p2 <- p + labs(title = tr, subtitle = NULL) +
      theme(plot.title = element_text(hjust = 0.5, face = "bold"))
    p <- p2
    NULL
  }, silent = TRUE)
  
  if(inherits(p_try, "try-error") || !inherits(p, "ggplot")) {
    p_grob <- ggpubr::as_ggplot(p)
    p <- p_grob + labs(title = tr, subtitle = NULL) +
      theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  }
  
  mps_list[[tr]] <- p2}

mps_panel <- ggarrange(
  mps_list[["GY"]],  mps_list[["OHSW"]], mps_list[["PN"]],
  mps_list[["PH"]],  mps_list[["NDF"]],  mps_list[["NDM"]],
  ncol = 3, nrow = 2,
  common.legend = TRUE, legend = "bottom",
  labels = c("A","B","C","D","E","F"),
  font.label = list(size = 10, face = "bold")
)

mps_panel <- annotate_figure(
  mps_panel,
  top = text_grob("Mean Performance vs Stability", face = "bold", size = 16)
)

mps_panel
ggsave("Mean Performance vs Stability_LB.pdf", mps_panel, width = 10, height = 8, dpi = 600)
```

#### Discriminativeness vs. representativeness

```{r}
library(ggplot2)
library(ggpubr)

library(ggplot2)
library(ggpubr)

dir.create("Discriminativeness_vs_Representativeness", showWarnings = FALSE)

dr_list <- list()

for(tr in traits){

  gge4_tr <- gge(
    LB,
    Env,
    GEN,
    resp = tr,
    centering = "environment",
    scaling = "none",
    svp = "environment"
  )
  p4 <- plot(gge4_tr,
             type = 4,
             col.env = "orange",
             col.gen = "purple",
             size.text.gen = 2,
             size.text.env = 4 ,
             repel = TRUE,
             repulsion = 2,
             col.line = "orange",
             size.line = 0.5,
             col.circle = "black",
             plot_theme = theme_metan_minimal())
  p_try <- try({
    p4_mod <- p4 +
      labs(title = tr, subtitle = NULL) +
      theme(plot.title = element_text(hjust = 0.5, face = "bold"))
    p4 <- p4_mod
    NULL
  }, silent = TRUE)
  if(inherits(p_try, "try-error") || !inherits(p4, "ggplot")) {
    p_grob <- ggpubr::as_ggplot(p4)
    p4 <- p_grob +
      labs(title = tr, subtitle = NULL) +
      theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  }
  dr_list[[tr]] <- p4
}
dr_panel <- ggarrange(
  dr_list[["GY"]],  dr_list[["OHSW"]], dr_list[["PN"]],
  dr_list[["PH"]],  dr_list[["NDF"]],  dr_list[["NDM"]],
  ncol = 3, nrow = 2,
  common.legend = TRUE, legend = "bottom",
  labels = c("A","B","C","D","E","F"),
  font.label = list(size = 10, face = "bold")
)

dr_panel <- annotate_figure(
  dr_panel,
  top = text_grob("Discriminativeness vs Representativeness", 
                  face = "bold", size = 16)
)

dr_panel
ggsave("Discriminativeness vs Representativeness_LB.pdf", dr_panel, width = 10, height = 8, dpi = 600)
```
