---
title: "Leveraging GxE Interaction to Optimize Multi-Trait Selection in Lima Bean"
date: "`r Sys.Date()`"
author: "João Paulo Silva Pavan"
format: html
editor: visual
---

```{r}
library(reshape2)
library(ggplot2)
library(openxlsx)
library(asreml)
library(dplyr)
library(tibble)
library(ggplot2)
library(metan)
library(writexl)
```

## Lima Bean Data

-   3 Locations (Brazil) - Piracicaba - SP, Teresina - PI, Tianguá - CE,

-   40 Lima Bean Lines

-   Randomized Complete Block Design (RCBD) - 3 Repetitions

-   Plot usable area: 5m²

-   Traits:

    |          Trait           | Acronym | Unit  |
    |:------------------------:|:-------:|:-----:|
    |       Grain Yield        |   GY    | kg/ha |
    | Number days to flowering |   NDF   | days  |
    | Number days to maturity  |   NDM   | days  |
    | One Hundred Seed Weight  |  OHSW   |   g   |
    |       Plant Height       |   PH    |  cm   |
    |        Pod Length        |   PL    |  mm   |
    |        Pod Number        |   PN    |  mm   |
    |      Pod Thickness       |   PT    |  mm   |
    |        Pod Width         |   PW    |  mm   |
    |       Seed Length        |   SL    |  mm   |
    |      Seed Thickness      |   ST    |  mm   |
    |        Seed Width        |   SW    |  mm   |

```{r}
LB = read.xlsx("C://Users//joaop//Desktop//fava//LimaBean.xlsx")
LB$Block    <- as.factor(LB$Block)
LB$Env    <- as.factor(LB$Env)
LB$Genotype <- as.factor(LB$Genotype)
LB$GEN<- as.factor(LB$GEN)
str(LB)
```

```{r}
hist(LB$NDF)
hist(LB$NDM)
hist(LB$PH)
hist(LB$PL)
hist(LB$PW)
hist(LB$PT)
hist(LB$SL)
hist(LB$SW)
hist(LB$ST)
hist(LB$PN)
hist(LB$GY)
hist(LB$OHSW)
```

## anova

```{r}
aov = anova_joint(LB,env = Env, gen = GEN, rep = Block,resp = everything())
```

## Mixed-effect models

```{r}
mixedmodel <- gamem_met(LB,env = Env, gen = GEN, rep = Block,random = "all",resp = everything())
```

#### LRT

```{r}
GYLRT = mixedmodel$GY$LRT
OHSWLRT = mixedmodel$OHSW$LRT 
NDFLRT = mixedmodel$NDF$LRT
NDMLRT = mixedmodel$NDM$LRT 
PHLRT = mixedmodel$PH$LRT
PNLRT = mixedmodel$PN$LRT
LRT = list(
  GY_LRT   = GYLRT,
  OHSW_LRT = OHSWLRT,
  NDF_LRT  = NDFLRT,
  NDM_LRT  = NDMLRT,
  PH_LRT   = PHLRT,
  PN_LRT   = PNLRT
)
LRT
```

```{r}
write_xlsx(LRT, "LRT_LB.xlsx")
```

#### BLUPs

```{r}
BLUPgen = list(
  GY_BLUP   = mixedmodel$GY$BLUPgen,
  OHSW_BLUP = mixedmodel$OHSW$BLUPgen,
  NDF_BLUP  = mixedmodel$NDF$BLUPgen,
  NDM_BLUP  = mixedmodel$NDM$BLUPgen,
  PH_BLUP   = mixedmodel$PH$BLUPgen,
  PN_BLUP   = mixedmodel$PN$BLUPgen
)
```

```{r}
write_xlsx(BLUPgen, "BLUPgen_LB.xlsx")
```

```{r}
library(dplyr)
library(tibble)

ebvMat <- data.frame(
  GEN = mixedmodel$GY$BLUPgen$GEN,
  GY   = mixedmodel$GY$BLUPgen$Predicted,
  OHSW = mixedmodel$OHSW$BLUPgen$Predicted,
  NDF  = mixedmodel$NDF$BLUPgen$Predicted,
  NDM  = mixedmodel$NDM$BLUPgen$Predicted,
  PH   = mixedmodel$PH$BLUPgen$Predicted,
  PN   = mixedmodel$PN$BLUPgen$Predicted
) |>
  column_to_rownames("GEN") |>
  as.matrix()
dim(ebvMat)
head(ebvMat)
```

```{r}
trait_units <- c(
  GY   = "kg/ha",   # kg·ha⁻¹
  OHSW = "g",
  PH   = "cm",
  PN   = "number",
  NDF  = "number",
  NDM  = "number"
)
plot_trait_heatmap <- function(x, trait_name) {
  
  gen_order <- sprintf("L%02d", 1:40)
  
  df <- x[[trait_name]]$BLUPgen %>%
    filter(GEN %in% gen_order) %>%
    mutate(
      GEN = factor(GEN, levels = gen_order),
      Trait = trait_name
    )
  
  ggplot(df, aes(x = GEN, y = Trait, fill = Predicted)) +
    geom_tile() +
    scale_fill_gradientn(
      colours = c("orange", "#E37383", "purple"),
      name = bquote(hat(y)~"("~.(trait_units[[trait_name]])~")")
    ) +
    labs(x = NULL, y = NULL) +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, face = "bold"),
      axis.text.y = element_text(angle = 90, face = "bold"),
      legend.position = "right",
      panel.grid = element_blank()
    )
}
traits <- c("GY", "OHSW", "PH", "PN", "NDF", "NDM")

BLUP_heatmaps <- lapply(traits, function(trait) {
  plot_trait_heatmap(mixedmodel, trait)
})
```

```{r}
blups_heatmap_panel <- ggarrange(
  plotlist = BLUP_heatmaps,
  ncol = 2, nrow = 3,
  labels = c("A", "B", "C", "D", "E", "F"), font.label = list(size = 10, face = "bold"),
  align = "v"
)
blups_heatmap_panel
```

```{r}
ggsave(
  "BLUPs_LB.pdf",
  blups_heatmap_panel,
  width = 14,
  height = 12,
  dpi = 600
)
```

#### Correlation

```{r}
corrplot = plot(
  corr_coef(LB, NDF, NDM, PH, PN, GY, OHSW),
  type = "upper",
  diag = FALSE,
  reorder = T,
  signif = c("stars", "pval"),
  show = c("all", "signif"),
  p_val = 0.05,
  caption = TRUE,
  digits.cor = 2,
  digits.pval = 3,
  col.low = "orange",
  col.mid = "white",
  col.high = "purple",
  lab.x.position = "top",
  lab.y.position = "left",   
  legend.position = "right",
  legend.title = "Pearson's\nCorrelation",
  size.text.cor = 5,
  size.text.signif = 5,
  size.text.lab = 14
)
corrplot
```

```{r}
ggsave("Correlation.pdf",corrplot, width = 7, height = 6, dpi = 600)
```

#### Genotype by trait biplot

```{r}
gtb1 = gtb(LB, GEN, resp = c(NDF, NDM, PH, PL, PW, 
             PN, GY, OHSW))

gtb1plot = plot(gtb1,
     col.env = "orange",
            col.gen = "purple",
            size.text.gen = 5,
            size.text.env = 8, repel = T,
            repulsion = 1,
            col.line = "orange",
            plot_theme = theme_metan_minimal())
```

```{r}
ggsave("Genotypebytraitbiplot_LB.pdf",gtb1plot, width = 8, height = 8, dpi = 600)
```

#### Which-won-Where

```{r}
library(ggpubr)
library(ggplot2)

www <- list()

for(tr in traits){
  gge3_tr <- gge(LB, Env, GEN, resp = tr, svp = "symmetrical")
  
  p <- plot(gge3_tr, type = 3,
            col.env = "orange",
            col.gen = "purple",
            arrow.color = "purple",
            size.text.gen = 1,
            size.text.env = 4,
            size.shape.win = 4,
            large_label = 1,
            repel = TRUE,
            size.text.win = 4,
            repulsion = 0.5,
            col.line = "orange",
            plot_theme = theme_metan_minimal())
  
  p_try <- try({
    p2 <- p + labs(title = tr, subtitle = NULL) +
            theme(plot.title = element_text(hjust = 0.5, face = "bold"))
    p <- p2
    NULL
  }, silent = TRUE)
  
  if(inherits(p_try, "try-error") || !inherits(p, "ggplot")) {
    p_grob <- ggpubr::as_ggplot(p)            
    p <- p_grob + labs(title = tr, subtitle = NULL) +
              theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  }
  
  www[[tr]] <- p
}

www_lb <- ggarrange(
  www[["GY"]],  www[["OHSW"]], www[["PN"]],
  www[["PH"]],  www[["NDF"]],  www[["NDM"]],
  ncol = 3, nrow = 2,
  common.legend = TRUE, legend = "bottom",
  labels = c("A","B","C","D","E","F"),
  font.label = list(size = 10, face = "bold"),
  align = "hv",
  hjust = -0.5,
  vjust = 1.5
)

www_lb <- annotate_figure(
  www_lb,
  top = text_grob("", size = 14),
  fig.lab = "",
  fig.lab.pos = "bottom"
)
www_lb
```

```{r}
ggsave("Which-won-Where.pdf", www_lb, width = 10, height = 14, dpi = 600)
```

#### Mean_performance vs. stability

```{r}
dir.create("Mean_performance_vs_stability", showWarnings = FALSE)

mps_list <- list()

for(tr in traits){
    gge2_tr <- gge(
    LB,
    Env,
    GEN,
    resp = tr,
    svp = "genotype"
  )
  
  # Plot type 2
  p <- plot(gge2_tr,
            type = 2,
            col.env = "orange",
            col.gen = "purple",
            size.text.gen = 2,
            size.text.env = 4,
            repel = TRUE,
            repulsion = 2,
            col.line = "orange",
            plot_theme = theme_bw())
  
  p_try <- try({
    p2 <- p + labs(title = tr, subtitle = NULL) +
      theme(plot.title = element_text(hjust = 0.5, face = "bold"))
    p <- p2
    NULL
  }, silent = TRUE)
  
  if(inherits(p_try, "try-error") || !inherits(p, "ggplot")) {
    p_grob <- ggpubr::as_ggplot(p)
    p <- p_grob + labs(title = tr, subtitle = NULL) +
      theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  }
  
  mps_list[[tr]] <- p2}

mps_panel <- ggarrange(
  mps_list[["GY"]],  mps_list[["OHSW"]], mps_list[["PN"]],
  mps_list[["PH"]],  mps_list[["NDF"]],  mps_list[["NDM"]],
  ncol = 3, nrow = 2,
  common.legend = TRUE, legend = "bottom",
  labels = c("A","B","C","D","E","F"),
  font.label = list(size = 10, face = "bold"),
  align = "hv"
)

mps_panel = annotate_figure(
  mps_panel,
  top = text_grob("", size = 14),
  fig.lab = "",
  fig.lab.pos = "bottom"
)
mps_panel
```

```{r}
ggsave("Mean Performance vs Stability_LB.pdf", mps_panel, width = 10, height = 14, dpi = 600)
```

#### Discriminativeness vs. representativeness

```{r}
library(ggplot2)
library(ggpubr)

dir.create("Discriminativeness_vs_Representativeness", showWarnings = FALSE)

dr_list <- list()

for(tr in traits){

  gge4_tr <- gge(
    LB,
    Env,
    GEN,
    resp = tr,
    centering = "environment",
    scaling = "none",
    svp = "environment"
  )
  p4 <- plot(gge4_tr,
             type = 4,
             col.env = "orange",
             col.gen = "purple",
             size.text.gen = 4,
             size.text.env = 4 ,
             repel = TRUE,
             repulsion = 2,
             col.line = "orange",
             size.line = 0.5,
             col.circle = "black",
             plot_theme = theme_metan_minimal())
  p_try <- try({
    p4_mod <- p4 +
      labs(title = tr, subtitle = NULL) +
      theme(plot.title = element_text(hjust = 0.5, face = "bold"))
    p4 <- p4_mod
    NULL
  }, silent = TRUE)
  if(inherits(p_try, "try-error") || !inherits(p4, "ggplot")) {
    p_grob <- ggpubr::as_ggplot(p4)
    p4 <- p_grob +
      labs(title = tr, subtitle = NULL) +
      theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  }
  dr_list[[tr]] <- p4
}
dr_panel <- ggarrange(
  dr_list[["GY"]],  dr_list[["OHSW"]], dr_list[["PN"]],
  dr_list[["PH"]],  dr_list[["NDF"]],  dr_list[["NDM"]],
  ncol = 3, nrow = 2,
  common.legend = TRUE, legend = "bottom",
  labels = c("A","B","C","D","E","F"),
  font.label = list(size = 10, face = "bold"), 
  align = "hv",
  hjust = -0.5,
  vjust = 1.5
  
)

dr_panel = annotate_figure(
  dr_panel,
  top = text_grob("", size = 14),
  fig.lab = "",
  fig.lab.pos = "bottom"
)
dr_panel
```

```{r}
ggsave("Discriminativeness vs Representativeness_LB.pdf", dr_panel, width = 10, height = 14, dpi = 600)
```

## Desire index

```{r}
# DESIRE INDEX FUNCTION
gainDiff <- function(ebvMat, b_ebv, dGain, nSel){ 
  
  take <- order(ebvMat %*% b_ebv, decreasing = TRUE)[1:nSel]
  
  mu_Sel <- colMeans(ebvMat[take, ])
  resp   <- mu_Sel - colMeans(ebvMat)
  resp   <- resp / sqrt(sum(resp^2))
  
  sum((resp - dGain)^2)
}
```

```{r}
ebvMat <- ebvMat[complete.cases(ebvMat), ]
#COV MATRIX -EBVS
library(AlphaSimR)

G_ebv <- AlphaSimR::popVar(ebvMat)
G_ebv
```

```{r}
#desired gains
desiredGain <- c(
  GY   =  2,
  OHSW =  0,
  NDF  = -1,
  NDM  = -1,
  PH   =  0,
  PN   =  1
)
desiredGain <- desiredGain / sqrt(sum(desiredGain^2))
sum(desiredGain^2)
```

```{r}
b_ebv <- solve(G_ebv) %*% desiredGain
b_ebv
```

```{r}
nSel <- 5
opt <- optim(
  par     = b_ebv,
  fn      = gainDiff,
  ebvMat  = ebvMat,
  dGain   = desiredGain,
  nSel    = nSel
)
opt_weights <- opt$par
opt_weights
```

```{r}
DesireIndex <- ebvMat %*% opt_weights

DesireIndex <- data.frame(
  GEN = rownames(ebvMat),
  DesireIndex = as.numeric(DesireIndex)
) |>
  arrange(desc(DesireIndex))

head(DesireIndex, 10)
```

```{r}
write.xlsx(DesireIndex, "Desireindex.xlsx")
```

```{r}
selected <- DesireIndex[1:5, ]
selected
```

```{r}
sel_mat <- ebvMat[selected$GEN, ]

expResp <- colMeans(sel_mat) - colMeans(ebvMat)
scaleExpResp <- expResp / sqrt(sum(expResp^2))

df_vect = cbind(
  Desired  = desiredGain,
  Achieved = scaleExpResp
)
```

```{r}
ggplot(df_vec, aes(x = Desired, y = Achieved)) +
  geom_point(size = 3, color = "purple") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_text(aes(label = Trait), nudge_x = 0.05, size = 5, color = "orange") +
  coord_equal(xlim = c(-1,1), ylim = c(-1,1)) +
  labs(
    x = "Desired Gain",
    y = "Achieved Gain"
  ) +
  theme_classic(base_size = 14)
```

```{r}
library(dplyr)
df <- dplyr::arrange(DesireIndex, dplyr::desc(DesireIndex)) |>
  dplyr::mutate(
    GEN = factor(GEN, levels = GEN),
    Selected = ifelse(GEN %in% selected$GEN, "Selected", "Not selected")
  )
```

```{r}
library(ggplot2)

desiredgainplot = ggplot(df, aes(x = GEN, y = DesireIndex, fill = Selected)) +
  geom_col(width = 0.9, color = "black", size = 0.2) + 
  coord_polar() +
  scale_fill_manual(
    values = c(
      "Selected"     = "#6A1B9A",
      "Not selected" = "orange"
    )
  ) +
  labs(face = "bold",
    fill  = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(face = "bold"),
    axis.title = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "bottom"
  )
```

```{r}
ggsave("desiredgainplot_LB.pdf", plot = desiredgainplot,width = 10, height = 8,dpi = 600)
```

### Table Selection Gain

```{r}
h2_df <- gmd(mixedmodel, "h2")
h2 <- h2_df$h2
names(h2) <- h2_df$VAR
traits_DI <- colnames(ebvMat)
h2_DI <- h2[traits_DI]
library(dplyr)

genetic_gain_table <- function(ebvMat, selected_gen, h2){

  traits <- colnames(ebvMat)

  res <- lapply(traits, function(tr){

    Xo <- mean(ebvMat[, tr], na.rm = TRUE)
    Xs <- mean(ebvMat[selected_gen, tr], na.rm = TRUE)

    SD  <- Xs - Xo
    SDp <- (SD / Xo) * 100

    SG  <- SD * h2[tr]
    SGp <- (SG / Xo) * 100

    tibble(
      Trait    = tr,
      `h²` = round(h2[tr], 2),
      Xo       = round(Xo, 2),
      Xs       = round(Xs, 2),
      SD       = round(SD, 2),
      `SD (%)` = round(SDp, 2),
      SG       = round(SG, 2),
      `SG (%)` = round(SGp, 2)
    )
  })

  bind_rows(res)
}
Table3_DesireIndex <- genetic_gain_table(
  ebvMat = ebvMat,
  selected_gen = selected$GEN,
  h2 = h2_DI
)

Table3_DesireIndex
```

```{r}
library(knitr)

tableDesire =kable(
  Table3_DesireIndex,
  caption = "Predicted genetic gain applying the Desire Index based on genotypic BLUPs."
)
```

```{r}
write.xlsx(Table3_DesireIndex, "Table3_DesireIndex.xlsx", rowNames = FALSE)
```

## Genotype by trait biplot

```{r}
gtb1 = gtb(LB, GEN, resp = c(NDF, NDM, PH, PN, GY, OHSW))

gtb1plot = plot(gtb1,
     col.env = "orange",
            col.gen = "purple",
            size.text.gen = 5,
            size.text.env = 8, repel = T,
            repulsion = 2,
            col.line = "orange",
            plot_theme = theme_metan())
gtb1plot

gtb1 = gtb(LB, GEN, resp = c(NDF, NDM, PH, PN, GY, OHSW))

gtb1plot = plot(gtb1,
     col.env = "orange",
            col.gen = "purple",
            size.text.gen = 5,
            size.text.env = 8, repel = T,
            repulsion = 2,
            col.line = "orange",
            plot_theme = theme_metan())
gtb1plot
```

### Panel Desire Gain Index + GYT Biplot

```{r}
gt_desired_panel <- ggarrange(gtb1plot, desiredgainplot,
  ncol = 2,
  common.legend = F, legend = "bottom",
  labels = c("A","B"),
  font.label = list(size = 10, face = "bold"), 
  align = "hv",
  hjust = -0.5,
  vjust = 1.5
)
gt_desired_panel
```

```{r}
ggsave("gt_desired_panel_LB.pdf", plot = gt_desired_panel, width = 12, height = 6,dpi = 600)
```

## Density plot

```{r}
if ("package:plyr" %in% search()) detach("package:plyr", unload = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggridges)
trait_names <- c(
  GY_BLUP   = "Grain Yield (GY)",
  OHSW_BLUP = "One Hundred Seed Weight (OHSW)",
  NDF_BLUP  = "Days to Flowering (NDF)",
  NDM_BLUP  = "Days to Maturity (NDM)",
  PH_BLUP   = "Plant Height (PH)",
  PN_BLUP   = "Pod Number (PN)"
)
df_long <- bind_rows(
  lapply(names(BLUPgen), function(tr) {
    BLUPgen[[tr]] %>%
      dplyr::mutate(
        Trait  = trait_names[tr],
        Status = ifelse(GEN %in% selected$GEN,
                        "Selected", "Not Selected")
      ) %>%
      select(GEN, Trait, Status, Predicted)
  })
)
df_long$Trait <- factor(
  df_long$Trait,
  levels = rev(trait_names)
)
```

```{r}
plot_density_SG <- function(data_trait, selected, xlab, bw = NULL) {
  
  df <- data_trait %>%
    dplyr::mutate(
      Status = factor(
        ifelse(GEN %in% selected$GEN, "Selected", "Not Selected"),
        levels = c("Not Selected", "Selected")
      )
    )
  
  ggplot(df, aes(x = Predicted, fill = Status)) +
    geom_density(alpha = 0.6, color = "black", linewidth = 0.4, bw = bw) +
    
    scale_fill_manual(
      values = c("Not Selected" = "#FFE5B4",
                 "Selected"     = "#BFB4FF")
    ) +
    
    labs(
      x = xlab,   # nome do trait + unidade
      y = "Density",
      fill = NULL
    ) +
    
    theme_light(base_size = 12) +
    theme(
      axis.title.x = element_text(face = "bold"),
      axis.title.y = element_text(face = "bold"),
      legend.position = "bottom"
    )
}

library(patchwork)

pd1 <- plot_density_SG(BLUPgen$GY_BLUP, selected,
                      expression(bold("GY (kg ha"^{-1}*")")), bw = 100)

pd2 <- plot_density_SG(BLUPgen$OHSW_BLUP, selected,
                      expression(bold("OHSW (g)")), bw = 2.5)

pd3 <- plot_density_SG(BLUPgen$NDF_BLUP, selected,
                      expression(bold("NDF (days)")), bw = 3)

pd4 <- plot_density_SG(BLUPgen$NDM_BLUP, selected,
                      expression(bold("NDM (days)")), bw = 3.5)

pd5 <- plot_density_SG(BLUPgen$PH_BLUP, selected,
                      expression(bold("PH (cm)")), bw = 0.5)

pd6 <- plot_density_SG(BLUPgen$PN_BLUP, selected,
                      expression(bold("Pod Number")), bw = 65)
densplot = (pd1 | pd2 | pd3) /
(pd4 | pd5 | pd6) +
  plot_annotation(
    tag_levels = "A",
    theme = theme(
      plot.tag = element_text(
        face = "bold",
        size = 14
      )
    )
  ) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
densplot
```

```{r}
ggsave(plot = densplot, "densplot.pdf", width = 12, height = 10, dpi = 600)
```
