---
project:
  type: website
  output-dir: docs

title: "Leveraging G×E Interaction to Optimize Multi-Trait Selection in Lima Bean"
author: "João Paulo Silva Pavan"
date: today

website:
  title: "Lima Bean G×E Analysis"
  navbar:
    left:
      - text: Home
        href: LimaBean_.qmd

format:
  html:
    toc: true
    page-layout: full
  pdf:
    toc: true

editor: visual
editor_options:
  chunk_output_type: console
---

### Packages

```{r, message= FALSE, warning= FALSE}
library(reshape2)
library(ggplot2)
library(openxlsx)
library(asreml)
library(dplyr)
library(tibble)
library(metan)
library(writexl)
```

```{r, message= FALSE, warning= FALSE}
knitr::opts_chunk$set(echo = TRUE)
  knitr::opts_knit$set(root.dir = "../")
```

## Lima Bean Data

-   3 Locations (Brazil) - Piracicaba - SP, Teresina - PI, Tianguá - CE,

-   40 Lima Bean Lines

-   Randomized Complete Block Design (RCBD) - 3 Repetitions

-   Plot usable area: 5m²

-   Traits:

    |          Trait           | Acronym | Unit  |
    |:------------------------:|:-------:|:-----:|
    |       Grain Yield        |   GY    | kg/ha |
    | Number days to flowering |   NDF   | days  |
    | Number days to maturity  |   NDM   | days  |
    | One Hundred Seed Weight  |  OHSW   |   g   |
    |       Plant Height       |   PH    |  cm   |
    |        Pod Length        |   PL    |  mm   |
    |        Pod Number        |   PN    |  mm   |
    |      Pod Thickness       |   PT    |  mm   |
    |        Pod Width         |   PW    |  mm   |
    |       Seed Length        |   SL    |  mm   |
    |      Seed Thickness      |   ST    |  mm   |
    |        Seed Width        |   SW    |  mm   |

#### Load the data

```{r, message= FALSE, warning= FALSE}
LB = read.xlsx("Data//LimaBean.xlsx")
LB$Block    <- as.factor(LB$Block)
LB$Env    <- as.factor(LB$Env)
LB$Genotype <- as.factor(LB$Genotype)
LB$GEN<- as.factor(LB$GEN)
str(LB)
```

### Histograms

```{r, eval=FALSE}
hist(LB$NDF)
hist(LB$NDM)
hist(LB$PH)
hist(LB$PL)
hist(LB$PW)
hist(LB$PT)
hist(LB$SL)
hist(LB$SW)
hist(LB$ST)
hist(LB$PN)
hist(LB$GY)
hist(LB$OHSW)
```

## Anova

```{r, eval=FALSE, message= FALSE, warning= FALSE}
aov = anova_joint(LB,env = Env, gen = GEN, rep = Block,resp = everything())
```

## Single environment analyses

```{r warning=FALSE, message=FALSE}
data.list = split(LB, f = LB$Env)
vccomp=list()
herit=list()
j="Piracicaba"
for (j in names(data.list)) {

   x = droplevels(data.list[[j]])
  cat("====> Environment:", j, fill = TRUE)

st <- gamem(x,
             gen = GEN,
             rep =  Block,
             resp = everything())
st
vccomp[[j]]=get_model_data(st, "vcomp")
vccomp[[j]]$env=j
herit[[j]]=get_model_data(st, "h2")
herit[[j]]$env=j
}
```

### Variance components

```{r, message= FALSE, warning= FALSE}
custom_palette=viridis::turbo(n = 19)
vc=do.call(rbind,vccomp)
colnames(vc)
traits=colnames(vc[,2:13])
vc_long <- reshape2::melt(vc, measure.vars = traits, variable.name = "trait")

vc_long$effect =paste(vc_long$Group,vc_long$env, sep = "_")

vcp <- vc_long |> 
  ggplot(aes(x = Group, y = value, fill = effect)) + 
  geom_col(position = "stack", just = 0.5) + 
  theme(axis.text.x = element_text(angle = 90), 
        strip.text = element_text(face = "bold")) + 
  facet_wrap(~trait, scales = "free") +
  labs(x = "Effects", y = "Variance components", fill = "Effects") +
  scale_fill_manual(
    values = custom_palette,  
  )
vcp
```

```{r, eval=FALSE}
ggsave(plot=vcp,device = "pdf",filename ="Plots/varcomp.pdf",dpi = "retina",height = 6,width = 8)
ggsave(plot=vcp,device = "png",filename ="Plots/varcomp.png",dpi = "retina",height = 6,width = 8,bg = "white")
```

### Broad-sense Heritability

```{r, fig.width=12, fig.height=8, message= FALSE, warning= FALSE}
custom_palette=viridis::turbo(n = 19)
her=do.call(rbind,herit)
colnames(her)

heritp <- her |> 
  ggplot(aes(x = VAR, y = h2, fill = env)) + 
  geom_col(position = "stack", just = 0.5) + 
  theme(axis.text.x = element_text(angle = 90), 
        strip.text = element_text(face = "bold")) + 
  facet_wrap(~env, scales = "free") +
  labs(x = "Traits", y = "Broad-sense heritability", fill = "Environments") +
  scale_fill_manual(
    values = custom_palette)
heritp
```

```{r, eval=FALSE}
ggsave(plot=heritp,device = "pdf",filename ="Plots/herit.pdf",dpi = "retina",height = 6,width = 8)
ggsave(plot=heritp,device = "png",filename ="Plots/herit.png",dpi = "retina",height = 6,width = 8,bg = "white")
```

## Mixed-effect models

```{r, message=FALSE, warning=FALSE}
mixedmodel <- gamem_met(LB,env = Env, gen = GEN, rep = Block,random = "gen" ,resp = everything())
```

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(purrr)

ne <- length(unique(LB$Env))
nr <- length(unique(LB$Block))

varcomp <- map_dfr(names(mixedmodel), function(tr){

  if(!"random" %in% names(mixedmodel[[tr]]))
    return(NULL)
  vc <- mixedmodel[[tr]]$random
  tibble(
    Trait = tr,
    sigma_g  = vc$Variance[vc$Group == "GEN"],
    sigma_ge = vc$Variance[vc$Group == "GEN:ENV"],
    sigma_e  = vc$Variance[vc$Group == "Residual"]
  )
})
genpar <- get_model_data(mixedmodel) %>%
  pivot_longer(
    cols = -Parameters,
    names_to = "Trait",
    values_to = "Value") %>%
  pivot_wider(
    names_from = Parameters,
    values_from = Value)

herit_table <- varcomp %>%
  left_join(genpar, by = "Trait") %>%
  dplyr::mutate(

    H2_individual =
      sigma_g /
      (sigma_g + sigma_ge + sigma_e),

    H2_MET =
      sigma_g /
      (sigma_g +
         sigma_ge/ne +
         sigma_e/(ne*nr)),

    H2mg_metan = h2mg) %>%
  dplyr::select(
    Trait,
    sigma_g,
    sigma_ge,
    sigma_e,
    H2_individual,
    H2_MET,
    H2mg_metan,
    Accuracy,
    GEIr2,
    rge,
    CVg,
    CVr,
    `CV ratio`)
herit_table
```

### Broad-sense Heritability MET

```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=6}
herit_met=get_model_data(mixedmodel, "h2")

custom_palette=viridis::turbo(n = 12)
heritp2 <- herit_met |> 
  ggplot(aes(x = VAR, y = h2, fill=VAR)) + 
  geom_col(position = "stack", just = 0.5) + 
  theme(axis.text.x = element_text(angle = 90), 
        strip.text = element_text(face = "bold")) +
  labs(x = "Traits", y = "Broad-sense heritability",fill = "Traits") +
  scale_fill_manual(
    values = custom_palette,  
  )
heritp2
```

```{r, eval=FALSE}
ggsave(plot=heritp2,device = "pdf",filename ="Plots/herit_met.pdf",dpi = "retina",height = 6,width = 8)
ggsave(plot=heritp2,device = "png",filename ="Plots/herit_met.png",dpi = "retina",height = 6,width = 8,bg = "white")
```

### Culllis Heritability

```{r, message=FALSE, warning=FALSE}
traits <- c("PH", "NDF", "NDM", "OHSW", "PN", "GY")
library(sommer)
library(purrr)
library(dplyr)
calc_cullis <- function(tr){
  mod <- mmer(
    fixed  = as.formula(
      paste0(tr," ~ Env + Block:Env")
    ),
    random = ~ GEN + GEN:Env,
    rcov   = ~ units,
    data   = LB
  )
  vg <- mod$sigma$GEN
  C22 <- mod$PevU$GEN[[tr]]
  n_g <- nrow(C22)
  trC22 <- sum(diag(C22))

  av2 <- 2/n_g *
    (trC22 -
      (sum(C22)-trC22)/(n_g-1))

  H2 <- 1 - av2/(2*vg)

  tibble(
    Trait = tr,
    Cullis_H2 = as.numeric(H2))

}
Cullis_all <- map_dfr(
  traits,
  calc_cullis)
Cullis_all
```

### LRT

```{r}
GYLRT = mixedmodel$GY$LRT
OHSWLRT = mixedmodel$OHSW$LRT 
NDFLRT = mixedmodel$NDF$LRT
NDMLRT = mixedmodel$NDM$LRT 
PHLRT = mixedmodel$PH$LRT
PNLRT = mixedmodel$PN$LRT
LRT = list(
  GY_LRT   = GYLRT,
  OHSW_LRT = OHSWLRT,
  NDF_LRT  = NDFLRT,
  NDM_LRT  = NDMLRT,
  PH_LRT   = PHLRT,
  PN_LRT   = PNLRT)
LRT
```

```{r, eval=FALSE}
write_xlsx(LRT, "LRT_LB.xlsx")
```

#### BLUPs

```{r}
BLUPgen = list(
  GY_BLUP   = mixedmodel$GY$BLUPgen,
  OHSW_BLUP = mixedmodel$OHSW$BLUPgen,
  NDF_BLUP  = mixedmodel$NDF$BLUPgen,
  NDM_BLUP  = mixedmodel$NDM$BLUPgen,
  PH_BLUP   = mixedmodel$PH$BLUPgen,
  PN_BLUP   = mixedmodel$PN$BLUPgen)
head(BLUPgen)
```

```{r, eval=FALSE}
write_xlsx(BLUPgen, "BLUPgen_LB.xlsx")
```

```{r, message=FALSE, warning=FALSE}
sommer_models <- setNames(
  map(traits, \(tr)
    mmer(
      fixed  = as.formula(
        paste0(tr, " ~ Env + Block:Env")
      ),
      random = ~ GEN + GEN:Env,
      rcov   = ~ units,
      data   = LB
    )
  ),
  traits)
sommer_blups <- map(
  traits,
  function(tr){

    data.frame(
      GEN = gsub("GEN", "", names(sommer_models[[tr]]$U$GEN[[tr]])),
      BLUP_sommer = as.numeric(
        sommer_models[[tr]]$U$GEN[[tr]]
      )
    )

  }
)
names(sommer_blups) <- traits
```

```{r}
#Compare BLUPs- gamem_met and mmer 
library(dplyr)

compare_blups <- map_dfr(
  traits,
  function(tr){

    gamem_blup <-
      mixedmodel[[tr]]$BLUPgen |>
      dplyr::select(GEN, BLUPg)

    sommer_blup <- sommer_blups[[tr]]

    comp <- left_join(
      gamem_blup,
      sommer_blup,
      by = "GEN")
    tibble(
      Trait = tr,
      Correlation =
        cor(
          comp$BLUPg,
          comp$BLUP_sommer
        ),
      Max_Difference =
        max(
          abs(
            comp$BLUPg -
            comp$BLUP_sommer
          )
        )
    )

  }
)
compare_blups
```

```{r}
library(dplyr)
library(tibble)

ebvMat <- data.frame(
  GEN = mixedmodel$GY$BLUPgen$GEN,
  GY   = mixedmodel$GY$BLUPgen$Predicted,
  OHSW = mixedmodel$OHSW$BLUPgen$Predicted,
  NDF  = mixedmodel$NDF$BLUPgen$Predicted,
  NDM  = mixedmodel$NDM$BLUPgen$Predicted,
  PH   = mixedmodel$PH$BLUPgen$Predicted,
  PN   = mixedmodel$PN$BLUPgen$Predicted
) |>
  column_to_rownames("GEN") |>
  as.matrix()
dim(ebvMat)
head(ebvMat)
save(ebvMat,file = "Data/ebvMat.rda")
```

```{r, eval=FALSE}
trait_units <- c(
  GY   = "kg/ha",   
  OHSW = "g",
  PH   = "cm",
  PN   = "number",
  NDF  = "number",
  NDM  = "number")
plot_trait_heatmap <- function(x, trait_name) {
  gen_order <- sprintf("L%02d", 1:40)
  df <- x[[trait_name]]$BLUPgen %>%
    filter(GEN %in% gen_order) %>%
    dplyr::mutate(
      GEN = factor(GEN, levels = gen_order),
      Trait = trait_name
    )
  ggplot(df, aes(x = GEN, y = Trait, fill = Predicted)) +
    geom_tile() +
    scale_fill_gradientn(
      colours = c("orange", "#E37383", "purple"),
      name = bquote(hat(y)~"("~.(trait_units[[trait_name]])~")")) +
    labs(x = NULL, y = NULL) +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, face = "bold"),
      axis.text.y = element_text(angle = 90, face = "bold"),
      legend.position = "right",
      panel.grid = element_blank())
}
traits <- c("GY", "OHSW", "PH", "PN", "NDF", "NDM")
BLUP_heatmaps <- lapply(traits, function(trait) {
  plot_trait_heatmap(mixedmodel, trait)
})
library(ggpubr)
blups_heatmap_panel <- ggarrange(
  plotlist = BLUP_heatmaps,
  ncol = 2, nrow = 3,
  labels = c("A", "B", "C", "D", "E", "F"), font.label = list(size = 10, face = "bold"),
  align = "v")
ggsave(
  "BLUPs_LB.pdf",
  blups_heatmap_panel,
  width = 14,
  height = 12,
  dpi = 600)
```

```{r, message= FALSE, warning= FALSE}
trait_units <- c(
  GY   = "kg/ha",   
  OHSW = "g",
  PH   = "cm",
  PN   = "number",
  NDF  = "number",
  NDM  = "number")
plot_trait_dot <- function(x, trait_name){
  gen_order <- sprintf("L%02d", 1:40)
  df <- x[[trait_name]]$BLUPgen %>%
    filter(GEN %in% gen_order) %>%
    dplyr::mutate(
      GEN = factor(GEN, levels = gen_order),
      Trait = trait_name)
  ggplot(df, aes(x = Trait, y = GEN)) +
    geom_point(
      aes(
        size  = round(abs(Predicted)),
        color = Predicted),
      alpha = 0.75) +
    scale_color_gradientn(
      colours = c("orange", "lightpink", "purple"),
      name = bquote(hat(y)~"("~.(trait_units[[trait_name]])~")"),
      guide = guide_colorbar(
    barheight = unit(0.3, "cm"),
    barwidth = unit(4, "cm"))) +

    scale_size(range = c(8,11), guide = "none") +

    labs(
      x = NULL,
      y = NULL) +

    theme_linedraw(base_size = 12) +
    theme(
      axis.text.x = element_text(
        angle = 0,
        hjust = 1,
        vjust = 1,
        face = "bold"),
      axis.text.y = element_text(face = "bold"),
      panel.grid = element_blank(),
      legend.position = "bottom",legend.title.position = "bottom",
      legend.text = element_text(size = 7.5, face = "bold"),
      legend.title = element_text(size = 8, face = "bold"))
}
traits <- c("GY", "OHSW", "PN", "PH", "NDF", "NDM")

BLUP_dots <- lapply(traits, function(tr){
  plot_trait_dot(mixedmodel, tr)
})

library(ggpubr)
#BLUP_dots
```

```{r,fig.width=14, fig.height=12, message= FALSE, warning= FALSE}
blups_d_panel <- ggarrange(
  plotlist = BLUP_dots,
  ncol = 6, nrow = 1,
  labels = c("A", "B", "C", "D", "E", "F"), vjust = 0.11,
  font.label = list(face = "bold"),label.y = 0.01,
  align = "v")
blups_d_panel
```

```{r, eval=FALSE}
ggsave(
  "BLUPs_d2_LB.pdf",
  blups_d_panel, height = 11, width = 13,
  dpi = 600)
```

#### Correlation

```{r, warning=FALSE, message=FALSE}
corrplot = plot(
  corr_coef(LB, NDF, NDM, PH, PN, GY, OHSW),
  type = "upper",
  diag = FALSE,
  reorder = T,
  signif = c("stars", "pval"),
  show = c("all", "signif"),
  p_val = 0.05,
  caption = TRUE,
  digits.cor = 2,
  digits.pval = 3,
  col.low = "orange",
  col.mid = "white",
  col.high = "purple",
  lab.x.position = "top",
  lab.y.position = "left",   
  legend.position = "right",
  legend.title = "Pearson's\nCorrelation",
  size.text.cor = 5,
  size.text.signif = 5,
  size.text.lab = 14)
```

```{r, eval=FALSE}
ggsave("Correlation.pdf",corrplot, width = 7, height = 6, dpi = 600)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE, echo=FALSE, fig.width=14, fig.height=10}
library(tidyverse)
library(igraph)
library(ggraph)
library(grid)
corr_df <- data.frame(
  Trait1 = c(
    "NDF","NDF","NDF","NDF","NDF",
    "NDM","NDM","NDM","NDM",
    "PH","PH","PH",
    "PN","PN",
    "GY"),
  Trait2 = c(
    "NDM","PH","PN","GY","OHSW",
    "PH","PN","GY","OHSW",
    "PN","GY","OHSW",
    "GY","OHSW",
    "OHSW"),
  Correlation = c(
    0.587, 0.205,-0.230,-0.264,-0.129,
    0.224,-0.214,-0.262,-0.131,
    0.160, 0.177, 0.413,
    0.861, 0.123,
    0.267))
library(igraph)
library(ggraph)
library(ggplot2)
library(grid)
# Graph
g <- graph_from_data_frame(corr_df, directed = FALSE)
# Plot
network_plot <- ggraph(g, layout = "fr") +
  geom_edge_link(
    aes(
      color = Correlation,
      width = abs(Correlation)
    ),
    alpha = 0.85
  ) +
  geom_node_point(size = 6, color = "black") +
  geom_node_text(
    aes(label = name),
    size = 5,
    fontface = "bold",
    repel = TRUE) +
  scale_edge_color_gradient2(
    low = "orange",
    mid = "lightpink",
    high = "purple4",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  scale_edge_width(range = c(0.3, 2)) +
  labs(
    edge_color = "Pearson correlation",
    edge_width = "") +
  guides(
    edge_width = guide_legend(title = NULL)) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold", size = 12),
    legend.text = element_text(size = 10))
ggsave("network_plot.pdf",network_plot, width = 7, height = 5, dpi = 600)
```

```{r, message=FALSE, warning=FALSE,fig.width=8, fig.height=8}
library(circlize)
# Correlation matrix
corr_mat <- matrix(
  c(
    1.000, 0.587, 0.205,-0.230,-0.264,-0.129,
    0.587, 1.000, 0.224,-0.214,-0.262,-0.131,
    0.205, 0.224, 1.000, 0.160, 0.177, 0.413,
   -0.230,-0.214, 0.160, 1.000, 0.861, 0.123,
   -0.264,-0.262, 0.177, 0.861, 1.000, 0.267,
   -0.129,-0.131, 0.413, 0.123, 0.267, 1.000),
  nrow = 6,
  byrow = TRUE)
traits <- c(
  "NDF",
  "NDM",
  "PH",
  "PN",
  "GY",
  "OHSW")
rownames(corr_mat) <- traits
colnames(corr_mat) <- traits
# Convert matrix to links
corr_df <- data.frame()
for(i in 1:(nrow(corr_mat)-1)){
  for(j in (i+1):ncol(corr_mat)){
    corr_df <- rbind(
      corr_df,
      data.frame(
        from  = rownames(corr_mat)[i],
        to    = colnames(corr_mat)[j],
        value = corr_mat[i,j]))
  }
}
# Link colors
corr_df$link_col <- ifelse(
  corr_df$value > 0,
  "purple",
  "orange")
# Trait colors
grid_cols <- c(
  NDF  = "grey60",
  NDM  = "grey40",
  PH   = "grey60",
  PN   = "grey40",
  GY   = "grey60",
  OHSW = "grey40")
# Plot
circos.clear()
chordDiagram(
  x = corr_df[, c("from","to","value")],
  grid.col = grid_cols,
  col = corr_df$link_col,
  transparency = 0.80,
  directional = 0,
  # MUCH THINNER LINES
  link.lwd = 0.5 + abs(corr_df$value) * 2,
  annotationTrack = "grid",
  preAllocateTracks = list(
    track.height = 0.12))
# Trait labels
circos.track(
  track.index = 1,
  panel.fun = function(x,y){
    circos.text(
      CELL_META$xcenter,
      CELL_META$ylim[1],
      CELL_META$sector.index,
      facing = "clockwise",
      niceFacing = TRUE,
      adj = c(0,0.5),
      font = 2,
      cex = 1.3)
  },
  bg.border = NA
)
# Title
title(
  "Pearson Correlation Network",
  font.main = 2,
  cex.main = 1.4)
# Legends
legend(
  "topleft",
  legend = c(
    "Positive",
    "Negative"),
  col = c(
    "purple",
    "orange"),
  lwd = 4,
  title = "Correlation sign",
  bty = "n",
  cex = 1)
legend(
  "topright",
  legend = c(
    "|r| = 0.2",
    "|r| = 0.5",
    "|r| = 0.8"),
  lwd = c(
    0.5 + 0.2*2,
    0.5 + 0.5*2,
    0.5 + 0.8*2),
  col = "black",
  title = "Correlation strength",
  bty = "n",
  cex = 1)
```

```{r, eval=FALSE}
pdf(
  "Correlation_Chord_Diagram.pdf",
  width = 8,
  height = 8)
dev.off()
```

## GGE BIPLOT

### Which-Won-where

```{r, fig.width=14, fig.height=16, message= FALSE, warning= FALSE}
www <- list()

for(tr in traits){
  gge3_tr <- gge(LB, Env, GEN, resp = tr, svp = "symmetrical")
  
  p <- plot(gge3_tr,
            type = 3,
            col.env = "orange",
            col.gen = "purple",
            shape.gen = 21,
            shape.env = 23,
            col.alpha = 0.3,
            arrow.color = "purple",
            size.text.gen = 4,
            size.text.win = 8,
            size.text.env = 8,
            size.shape = 5,
            size.shape.win = 10,
            large_label = 10,
            repel = TRUE,
            max_overlaps = 40, 
            repulsion =8,
            col.line = "orange",
            plot_theme = theme_metan_minimal())
  
  p_try <- try({
    p2 <- p + labs(title = tr, subtitle = NULL) +
            theme(plot.title = element_text(hjust = 0.5, face = "bold"))
    p <- p2
    NULL
  }, silent = TRUE)
  
  if(inherits(p_try, "try-error") || !inherits(p, "ggplot")) {
    p_grob <- ggpubr::as_ggplot(p)            
    p <- p_grob + labs(title = tr, subtitle = NULL) +
              theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  }
  
  www[[tr]] <- p
}
www_lb <- ggarrange(
  www[["GY"]],  www[["OHSW"]], www[["PN"]],
  www[["PH"]],  www[["NDF"]],  www[["NDM"]],
  ncol = 2, nrow = 3,
  common.legend = TRUE, legend = "bottom",
  labels = c("A","B","C","D","E","F"),
  font.label = list(size = 12, face = "bold"),
  align = "hv",
  hjust = -0.5,
  vjust = 1.5)
www_lb
```

```{r, eval=FALSE}
ggsave("Which-won-Where2.pdf", www_lb,width = 14, height = 16, dpi = 600)
```

### Mean_performance vs. stability

```{r, message= FALSE, warning= FALSE}
dir.create("Mean_performance_vs_stability", showWarnings = FALSE)

mps_list <- list()

for(tr in traits){
    gge2_tr <- gge(
    LB,
    Env,
    GEN,
    resp = tr,
    svp = "genotype")
  # Plot type 2
  p <- plot(gge2_tr,
            type = 2,
            col.env = "orange",
            col.gen = "purple",
            shape.gen = 21,
            shape.env = 23,
            col.alpha = 0.3,
            arrow.color = "purple",
            size.text.gen = 4,
            size.text.env = 5,
            size.shape = 5,
            large_label = 1,
            repel = TRUE,
            max_overlaps = 10, 
            repulsion =40,
            col.line = "orange",
            axis_expand = 1.2,
            plot_theme = theme_metan_minimal())
  
  p_try <- try({
    p2 <- p + labs(title = tr, subtitle = NULL) +
      theme(plot.title = element_text(hjust = 0.5, face = "bold"))
    p <- p2
    NULL
  }, silent = TRUE)
  
  if(inherits(p_try, "try-error") || !inherits(p, "ggplot")) {
    p_grob <- ggpubr::as_ggplot(p)
    p <- p_grob + labs(title = tr, subtitle = NULL) +
      theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  }
  
  mps_list[[tr]] <- p2}

mps_panel <- ggarrange(
  mps_list[["GY"]],  mps_list[["OHSW"]], mps_list[["PN"]],
  mps_list[["PH"]],  mps_list[["NDF"]],  mps_list[["NDM"]],
  ncol = 2, nrow = 3,
  common.legend = TRUE, legend = "bottom",
  labels = c("A","B","C","D","E","F"),
  font.label = list(size = 10, face = "bold"),
  align = "hv")
mps_panel = annotate_figure(
  mps_panel,
  top = text_grob("", size = 14),
  fig.lab = "",
  fig.lab.pos = "bottom")
```

```{r, message= FALSE, warning= FALSE}
mps_GY <- gge(
  LB,
  Env,
  GEN,
  resp = "GY",
  svp = "genotype")

mpsp_GY <- plot(
  mps_GY,
  type = 2,
  col.env = "orange",
  col.gen = "purple",
  col.line = "grey50",
  size.text.gen = 4,
  size.text.env = 5,
  repel = TRUE,
  repulsion = 0,
  max_overlaps = 40,
  axis_expand = 1.1,
  plot_theme = theme_classic(base_size = 12)) +
  labs(title = NULL, subtitle = NULL)

mpsp_OHSW <- gge(
  LB,
  Env,
  GEN,
  resp = "OHSW",
  svp = "genotype")

mpspp_OHSW <- plot(
  mpsp_OHSW,
  type = 2,
  col.env = "orange",
  col.gen = "purple",
  col.line = "grey50",
  size.text.gen = 4,
  size.text.env = 5,
  repel = TRUE,
  repulsion = 20,
  max_overlaps = 30,
  axis_expand = 1.08,
  plot_theme = theme_classic(base_size = 14)) +
  labs(title = NULL, subtitle = NULL)

mpsp_PN <- gge(
  LB,
  Env,
  GEN,
  resp = "PN",
  svp = "genotype")

mpspp_PN <- plot(
  mpsp_PN,
  type = 2,
  col.env = "orange",
  col.gen = "purple",
  col.line = "grey50",
  size.text.gen = 4,
  size.text.env = 5,
  repel = TRUE,
  repulsion = 20,
  max_overlaps = 30,
  axis_expand = 1.1,
  plot_theme = theme_classic(base_size = 14)) +
  labs(title = NULL, subtitle = NULL)

mpsp_PH <- gge(
  LB,
  Env,
  GEN,
  resp = "PH",
  svp = "genotype")

mpspp_PH <- plot(
  mpsp_PH,
  type = 2,
  col.env = "orange",
  col.gen = "purple",
  col.line = "grey50",
  size.text.gen = 4,
  size.text.env = 5,
  repel = TRUE,
  repulsion = 20,
  max_overlaps = 30,
  axis_expand = 1.05,
  plot_theme = theme_classic(base_size = 14)) +
  labs(title = NULL, subtitle = NULL)

mpsp_NDF <- gge(
  LB,
  Env,
  GEN,
  resp = "NDF",
  svp = "genotype")

mpspp_NDF <- plot(
  mpsp_NDF,
  type = 2,
  col.env = "orange",
  col.gen = "purple",
  col.line = "grey50",
  size.text.gen = 4,
  size.text.env = 5,
  repel = TRUE,
  repulsion = 20,
  max_overlaps = 30,
  axis_expand = 1.1,
  plot_theme = theme_classic(base_size = 14)) +
  labs(title = NULL, subtitle = NULL)

mpsp_NDM <- gge(
  LB,
  Env,
  GEN,
  resp = "NDM",
  svp = "genotype")

mpspp_NDM <- plot(
  mpsp_NDM,
  type = 2,
  col.env = "orange",
  col.gen = "purple",
  col.line = "grey50",
  size.text.gen = 4,
  size.text.env = 5,
  repel = TRUE,
  repulsion = 20,
  max_overlaps = 30,
  axis_expand = 1.07,
  plot_theme = theme_classic(base_size = 14)) +
  labs(title = NULL, subtitle = NULL)
```

```{r, message= FALSE, warning= FALSE}
mpsp_GY <- mpsp_GY +
  ggtitle("GY") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

mpspp_OHSW <- mpspp_OHSW +
  ggtitle("OHSW") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

mpspp_PN <- mpspp_PN +
  ggtitle("PN") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

mpspp_PH <- mpspp_PH +
  ggtitle("PH") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

mpspp_NDF <- mpspp_NDF +
  ggtitle("NDF") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

mpspp_NDM <- mpspp_NDM +
  ggtitle("NDM") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

```{r,fig.width=14, fig.height=16, message= FALSE, warning= FALSE}
common_theme <- theme(
  legend.position = "bottom",
  panel.grid = element_blank(),
  plot.title = element_text(
    hjust = 0.5,
    face = "bold",
    size = 12
  )
)
mpsp_GY     <- mpsp_GY + common_theme
mpspp_OHSW  <- mpspp_OHSW + common_theme
mpspp_PN    <- mpspp_PN + common_theme
mpspp_PH    <- mpspp_PH + common_theme
mpspp_NDF   <- mpspp_NDF + common_theme
mpspp_NDM   <- mpspp_NDM + common_theme
mps_panel <- ggarrange(
  mpsp_GY,
  mpspp_OHSW,
  mpspp_PN,
  mpspp_PH,
  mpspp_NDF,
  mpspp_NDM,
  ncol = 2,
  nrow = 3,
  common.legend = TRUE,
  legend = "bottom",
  labels = c("A","B","C","D","E","F"),
  font.label = list(
    size = 12,
    face = "bold"
  )
)
mps_panel
```

```{r, eval=FALSE}
ggsave("Mean Performance vs Stability_LB2.pdf", mps_panel, width = 14, height = 16, dpi = 600)
```

### Discriminativeness vs. representativeness

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(ggpubr)

dir.create("Discriminativeness_vs_Representativeness", showWarnings = FALSE)

dr_list <- list()

for(tr in traits){

  gge4_tr <- gge(
    LB,
    Env,
    GEN,
    resp = tr,
    centering = "environment",
    scaling = "none",
    svp = "environment"
  )
  
  p4 <- plot(gge4_tr,
             type = 4,
             col.env = "orange",
             col.gen = "purple",
             size.text.gen = 4,
             size.text.env = 4 ,
             repel = TRUE,
             repulsion = 10,
             col.line = "orange",
             size.line = 0.5,
             col.circle = "grey",
             col.alpha = 0.4,
             max_overlaps = 20,
             plot_theme = theme_metan_minimal())
  p_try <- try({
    p4_mod <- p4 +
      labs(title = tr, subtitle = NULL) +
      theme(plot.title = element_text(hjust = 0.5, face = "bold"))
    p4 <- p4_mod
    NULL
  }, silent = TRUE)
  if(inherits(p_try, "try-error") || !inherits(p4, "ggplot")) {
    p_grob <- ggpubr::as_ggplot(p4)
    p4 <- p_grob +
      labs(title = tr, subtitle = NULL) +
      theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  }
  dr_list[[tr]] <- p4
}
dr_panel <- ggarrange(
  dr_list[["GY"]],  dr_list[["OHSW"]], dr_list[["PN"]],
  dr_list[["PH"]],  dr_list[["NDF"]],  dr_list[["NDM"]],
  ncol = 2, nrow = 3,
  common.legend = TRUE, legend = "bottom",
  labels = c("A","B","C","D","E","F"),
  font.label = list(size = 10, face = "bold"), 
  align = "hv",
  hjust = -0.5,
  vjust = 1.5
)
dr_panel = annotate_figure(
  dr_panel,
  top = text_grob("", size = 14),
  fig.lab = "",
  fig.lab.pos = "bottom"
)
```

```{r, message= FALSE, warning= FALSE}
dr_GY <- gge(
  LB,
  Env,
  GEN,
  resp = "GY",
  centering = "environment",
  scaling = "none",
  svp = "environment"
)

drp_GY <- plot(
  dr_GY,
  type = 4,
  col.env = "orange",
  col.gen = "purple",
  size.text.gen = 3,
  size.text.env = 5,
  col.alpha = 0.2,
  repel = TRUE,
  repulsion = 5,
  max_overlaps = 20,
  col.line = "darkorange",
  plot_theme = theme_classic(base_size = 12)
) +
  labs(title = NULL, subtitle = NULL)

dr_OHSW <- gge(
  LB,
  Env,
  GEN,
  resp = "OHSW",
  centering = "environment",
  scaling = "none",
  svp = "environment"
)

drp_OHSW <- plot(
  dr_OHSW,
  type = 4,
  col.env = "orange",
  col.gen = "purple",
  size.text.gen = 3,
  size.text.env = 5,
  col.alpha = 0.2,
  repel = TRUE,
  repulsion = 5,
  max_overlaps = 20,
  col.line = "darkorange",
  plot_theme = theme_classic(base_size = 12)
) +
  labs(title = NULL, subtitle = NULL)

dr_PN <- gge(
  LB,
  Env,
  GEN,
  resp = "PN",
  centering = "environment",
  scaling = "none",
  svp = "environment"
)

drp_PN <- plot(
  dr_PN,
  type = 4,
  col.env = "orange",
  col.gen = "purple",
  size.text.gen = 3,
  size.text.env = 5,
  col.alpha = 0.2,
  repel = TRUE,
  repulsion = 5,
  max_overlaps = 10,
  col.line = "darkorange",
  plot_theme = theme_classic(base_size = 12)
) +
  labs(title = NULL, subtitle = NULL)

dr_PH<- gge(
  LB,
  Env,
  GEN,
  resp = "PH",
  centering = "environment",
  scaling = "none",
  svp = "environment"
)

drp_PH <- plot(
  dr_PH,
  type = 4,
  col.env = "orange",
  col.gen = "purple",
  size.text.gen = 3,
  size.text.env = 5,
  col.alpha = 0.2,
  repel = TRUE,
  repulsion = 5,
  max_overlaps = 20,
  col.line = "darkorange",
  plot_theme = theme_classic(base_size = 12)
) +
  labs(title = NULL, subtitle = NULL)

dr_NDF<- gge(
  LB,
  Env,
  GEN,
  resp = "NDF",
  centering = "environment",
  scaling = "none",
  svp = "environment"
)

drp_NDF <- plot(
  dr_NDF,
  type = 4,
  col.env = "orange",
  col.gen = "purple",
  size.text.gen = 3,
  size.text.env = 5,
  col.alpha = 0.2,
  repel = TRUE,
  repulsion = 5,
  max_overlaps = 20,
  col.line = "darkorange",
  plot_theme = theme_classic(base_size = 12)
) +
  labs(title = NULL, subtitle = NULL)

dr_NDM<- gge(
  LB,
  Env,
  GEN,
  resp = "NDM",
  centering = "environment",
  scaling = "none",
  svp = "environment"
)

drp_NDM <- plot(
  dr_NDM,
  type = 4,
  col.env = "orange",
  col.gen = "purple",
  size.text.gen = 3,
  size.text.env = 5,
  col.alpha = 0.2,
  repel = TRUE,
  repulsion = 5,
  max_overlaps = 10,
  col.line = "darkorange",
  plot_theme = theme_classic(base_size = 12)
) +
  labs(title = NULL, subtitle = NULL)
```

```{r, fig.width=14, fig.height=16, message= FALSE, warning= FALSE}
drp_GY <- drp_GY +
  ggtitle("GY") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

drp_OHSW <- drp_OHSW +
  ggtitle("OHSW") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

drp_PN <- drp_PN +
  ggtitle("PN") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

drp_PH <- drp_PH +
  ggtitle("PH") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

drp_NDF <- drp_NDF +
  ggtitle("NDF") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

drp_NDM <- drp_NDM +
  ggtitle("NDM") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
common_theme <- theme(
  legend.position = "bottom",
  panel.grid = element_blank(),
  plot.title = element_text(
    hjust = 0.5,
    face = "bold",
    size = 12))
drp_GY    <- drp_GY    + common_theme
drp_OHSW  <- drp_OHSW  + common_theme
drp_PN    <- drp_PN    + common_theme
drp_PH    <- drp_PH    + common_theme
drp_NDF   <- drp_NDF   + common_theme
drp_NDM   <- drp_NDM   + common_theme
dr_panel <- ggarrange(
  drp_GY,
  drp_OHSW,
  drp_PN,
  drp_PH,
  drp_NDF,
  drp_NDM,
  ncol = 2,
  nrow = 3,
  common.legend = TRUE,
  legend = "bottom",
  labels = c("A","B","C","D","E","F"),
  font.label = list(
    size = 12,
    face = "bold"
  )
)
dr_panel
```

```{r, eval=FALSE}
ggsave("Discriminativeness vs Representativeness_LB.pdf", dr_panel, width = 14, height = 16, dpi = 600)
```

# Simultaneous Selection

## Desire Gain Index

```{r}
# DESIRE INDEX FUNCTION
gainDiff <- function(ebvMat, b_ebv, dGain, nSel){ 
  take <- order(ebvMat %*% b_ebv, decreasing = TRUE)[1:nSel]
  mu_Sel <- colMeans(ebvMat[take, ])
  resp   <- mu_Sel - colMeans(ebvMat)
  resp   <- resp / sqrt(sum(resp^2))
  sum((resp - dGain)^2)
}
```

```{r}
ebvMat <- ebvMat[complete.cases(ebvMat), ]
#COV MATRIX -EBVS
library(AlphaSimR)
(G_ebv <- AlphaSimR::popVar(ebvMat))
```

```{r}
#desired gains
desiredGain <- c(
  GY   =  2,
  OHSW =  1,
  NDF  = -1,
  NDM  = -1,
  PH   =  0,
  PN   =  1
)
desiredGain <- desiredGain / sqrt(sum(desiredGain^2))
sum(desiredGain^2)
```

```{r}
b_ebv <- solve(G_ebv) %*% desiredGain
```

```{r}
nSel <- 5
opt <- optim(
  par     = b_ebv,
  fn      = gainDiff,
  ebvMat  = ebvMat,
  dGain   = desiredGain,
  nSel    = nSel
)
(opt_weights <- opt$par)
```

```{r}
DesireIndex <- ebvMat %*% opt_weights
DesireIndex <- data.frame(
  GEN = rownames(ebvMat),
  DesireIndex = as.numeric(DesireIndex)) |>
  arrange(desc(DesireIndex))
head(DesireIndex, 10)
```

```{r, eval=FALSE}
write.xlsx(DesireIndex, "Desireindex.xlsx")
```

```{r}
(selected <- DesireIndex[1:5, ])
```

```{r}
sel_mat <- ebvMat[selected$GEN, ]

expResp <- colMeans(sel_mat) - colMeans(ebvMat)
scaleExpResp <- expResp / sqrt(sum(expResp^2))

df_vect = cbind(
  Desired  = desiredGain,
  Achieved = scaleExpResp)
```

```{r, eval=FALSE, echo=FALSE, message= FALSE, warning= FALSE}
ggplot(df_vect, aes(x = Desired, y = Achieved)) +
  geom_point(size = 3, color = "purple") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_text(aes(label = ""), nudge_x = 0.05, size = 5, color = "orange") +
  coord_equal(xlim = c(-1,1), ylim = c(-1,1)) +
  labs(
    x = "Desired Gain",
    y = "Achieved Gain"
  ) +
  theme_classic(base_size = 14)
```

```{r}
library(dplyr)
df <- dplyr::arrange(DesireIndex, dplyr::desc(DesireIndex)) |>
  dplyr::mutate(
    GEN = factor(GEN, levels = GEN),
    Selected = ifelse(GEN %in% selected$GEN, "Selected", "Not selected"))
```

```{r, message= FALSE, warning= FALSE}
library(ggplot2)

desiredgainplot = ggplot(df, aes(x = GEN, y = DesireIndex, fill = Selected)) +
  geom_col(width = 0.9, color = "black", size = 0.2) + 
  coord_polar() +
  scale_fill_manual(
    values = c(
      "Selected"     = "#6A1B9A",
      "Not selected" = "orange")) +
  labs(face = "bold",
    fill  = NULL) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(face = "bold"),
    axis.title = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "bottom")
```

```{r, eval=FALSE}
ggsave("desiredgainplot_LB.pdf", plot = desiredgainplot,width = 10, height = 8,dpi = 600)
```

```{r}
h2_df <- gmd(mixedmodel, "h2")
h2 <- h2_df$h2
names(h2) <- h2_df$VAR
traits_DI <- colnames(ebvMat)
h2_DI <- h2[traits_DI]
library(dplyr)

genetic_gain_table <- function(ebvMat, selected_gen, h2){

  traits <- colnames(ebvMat)

  res <- lapply(traits, function(tr){

    Xo <- mean(ebvMat[, tr], na.rm = TRUE)
    Xs <- mean(ebvMat[selected_gen, tr], na.rm = TRUE)

    SD  <- Xs - Xo
    SDp <- (SD / Xo) * 100

    SG  <- SD * h2[tr]
    SGp <- (SG / Xo) * 100

    tibble(
      Trait    = tr,
      `h²` = round(h2[tr], 2),
      Xo       = round(Xo, 2),
      Xs       = round(Xs, 2),
      SD       = round(SD, 2),
      `SD (%)` = round(SDp, 2),
      SG       = round(SG, 2),
      `SG (%)` = round(SGp, 2)
    )
  })

  bind_rows(res)
}
Table3_DesireIndex <- genetic_gain_table(
  ebvMat = ebvMat,
  selected_gen = selected$GEN,
  h2 = h2_DI)
```

```{r}
h2_cullis <- Cullis_all$Cullis_H2
names(h2_cullis) <- Cullis_all$Trait
h2_cullis
Table3_DesireIndex_Cullis <- genetic_gain_table(
  ebvMat = ebvMat,
  selected_gen = selected$GEN,
  h2 = h2_cullis)
Table3_DesireIndex_Cullis
```

```{r}
Table_Broad <- genetic_gain_table(
  ebvMat,
  selected$GEN,
  h2_DI) %>%
  dplyr::rename(
    H2_Broad = `h²`,
    SG_Broad = SG,
    `SG_Broad (%)` = `SG (%)`)

Table_Cullis <- genetic_gain_table(
  ebvMat,
  selected$GEN,
  h2_cullis) %>%
  dplyr::rename(
    H2_Cullis = `h²`,
    SG_Cullis = SG,
    `SG_Cullis (%)` = `SG (%)`)

CompareGain <- Table_Broad %>%
  dplyr::select(Trait, H2_Broad, SG_Broad, `SG_Broad (%)`) %>%
  left_join(
    Table_Cullis %>%
      dplyr::select(Trait, H2_Cullis, SG_Cullis, `SG_Cullis (%)`),
    by = "Trait")
CompareGain
```

```{r}
library(knitr)
tableDesire =kable(Table3_DesireIndex)
```

```{r, eval=FALSE}
write.xlsx(Table3_DesireIndex, "Table3_DesireIndex.xlsx", rowNames = FALSE)
```

### Populations

```{r}
library(tibble)
populations <- tribble(
  ~Population, ~Cross, ~Male_origin, ~Female_origin, ~Derived_lines,
  "H25", "G25236 × BGP UFPI 628",
  "Buenos Aires, Argentina (CIAT)",
  "Piauí, Brazil (BGP-UFPI)",
  "L01–L16",

  "H46", "UC HASKELL × BGP UFPI 728",
  "California, USA (UC Davis)",
  "Piauí, Brazil (BGP-UFPI)",
  "L17–L30",

  "H50", "BGP UFPI 728 × BGP UFPI 628",
  "Piauí, Brazil (BGP-UFPI)",
  "Piauí, Brazil (BGP-UFPI)",
  "L31–L34",

  "H81", "BGP UFPI 628 × UC 92",
  "Piauí, Brazil (BGP-UFPI)",
  "California, USA (UC Davis)",
  "L35–L37",

  "H94", "BGP UFPI 728 × UC 92",
  "Piauí, Brazil (BGP-UFPI)",
  "California, USA (UC Davis)",
  "L38–L40")
populations
```

```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
pop_lines <- populations %>%
  rowwise() %>%
  dplyr::mutate(
    GEN = list(
      sprintf(
        "L%02d",
        seq(
          as.numeric(str_extract(Derived_lines, "(?<=L)\\d+")),
          as.numeric(str_extract(Derived_lines, "(?<=–L)\\d+"))
        )
      )
    )
  ) %>%
  unnest(GEN) %>%
  ungroup()
```

```{r}
DesireIndex_pop <- DesireIndex %>%
  left_join(
    pop_lines %>%
      dplyr::select(GEN, Population, Cross, Derived_lines),
    by = "GEN")
```

```{r}
pop_summary <- DesireIndex_pop %>%
  group_by(Population, Cross) %>%
  summarise(
    Mean_DI = mean(DesireIndex),
    .groups = "drop")

ggplot(
  pop_summary,
  aes(
    x = reorder(Population, Mean_DI),
    y = Mean_DI,
    fill = Population)) +
  geom_col() +
  coord_flip() +
  theme_minimal(base_size = 14)
```

```{r}
top5 <- selected$GEN
DesireIndex_pop <- DesireIndex_pop %>%
  dplyr::mutate(
    Elite = ifelse(GEN %in% top5, "Selected", "Not Selected"))
```

```{r,fig.width=14, fig.height=16, message= FALSE, warning= FALSE}
library(ggplot2)
library(ggrepel)

DGIPLOT = ggplot(
  DesireIndex_pop,
  aes(
    x = Population,
    y = DesireIndex
  )
) +

  geom_boxplot(
    aes(fill = Cross),
    width = 0.9,
    alpha = 0.13,
    linewidth = 0.5,
    outlier.shape = NA
  ) +

  # Points
  geom_jitter(
    aes(
      color = Elite,
      shape = Elite),
    width = 0.15,
    size = 3,
    alpha = 0.8) +
  # Labels for non-selected lines
  geom_text_repel(
    data = subset(DesireIndex_pop, Elite == "Not Selected"),
    aes(label = GEN),
    size = 4,
    fontface = "bold",
    alpha = 0.75,
    color = "black",
    max.overlaps = Inf,
    box.padding = 0,
    point.padding = 1,
    segment.alpha = 0) +
  # Labels for selected lines
  geom_text_repel(
    data = subset(DesireIndex_pop, Elite == "Selected"),
    aes(label = GEN),
    size = 6,
    fontface = "bold",
    alpha = 0.75,
    color = "purple",
    max.overlaps = Inf,
    box.padding = 0.3,
    point.padding = 1,
    segment.alpha = 0) +
  scale_shape_manual(
    values = c(
      "Not Selected" = 20,
      "Selected" = 20
    ),
  guide = "none") +
  scale_color_manual(
    values = c(
      "Not Selected" = "black",
      "Selected" = "purple"),
  guide = "none")  +
  scale_fill_manual(
  values = c(
    "G25236 × BGP UFPI 628"       = "#FF5E66",
    "UC HASKELL × BGP UFPI 728"   = "#FFF75E",
    "BGP UFPI 728 × BGP UFPI 628" = "#5EB7FF",
    "BGP UFPI 628 × UC 92"        = "#66FF5E",
    "BGP UFPI 728 × UC 92"        = "#FF5EB7"
  ),
  name = "Cross" )+
  labs(
    x = "Population",
    y = "Desire Gain Index",
    color = NULL,
    shape = NULL) +
  theme_classic(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(
      colour = "black",
      linewidth = 1),
    axis.title.x = element_text(
      face = "bold",
      size = 14),
    axis.title.y = element_text(
      face = "bold",
      size = 14),
    axis.text.x = element_text(
      face = "bold",
      size = 12),
    axis.text.y = element_text(
      face = "bold",
      size = 12),
    legend.text = element_text(
      face = "bold",
      size = 6),
    legend.position = c(0.55, 0.18)
  )
DGIPLOT
```

## Genotype by trait (GT) biplot

```{r, eval=FALSE, echo=FALSE}
gtb1 = gtb(LB, GEN, resp = c(NDF, NDM, PH, PN, GY, OHSW))
gtb1plot <- plot(
  gtb1,
  col.env = "orange",
  col.gen = "purple",
  size.text.gen = 4,
  size.text.env = 4,
  repel = TRUE,
  repulsion = 4,
  alpha = 0.5,
  col.line = "orange",
  print.sub = FALSE,
  plot_theme = theme_bw(base_size = 14) +
    theme(
      plot.title = element_blank(),
      plot.subtitle = element_blank(),
      panel.grid = element_blank(),
      legend.position = c(0.85, 0.2),
      legend.title = element_blank())) +
  annotate(
    "text",
    x = Inf, y = -Inf,
    label = "Scaling = 1, Centering = 2, SVP = 2",
    hjust = 1.1,
    vjust = -0.8,
    fontface = "bold",
    size = 3)
```

```{r, message= FALSE, warning= FALSE}
library(ggplot2)
gtb1 <- gtb(
  LB,
  GEN,
  resp = c(NDF, NDM, PH, PN, GY, OHSW))
gtb1plot <- plot(
  gtb1,
  col.env = "orange",
  col.gen = "purple",
  size.text.gen = 4,
  size.text.env = 4,
  repel = TRUE,
  repulsion = 4,
  col.line = "darkorange",
  print.sub = FALSE,
  plot_theme = theme_bw(base_size = 14) +
    theme(
      plot.title = element_blank(),
      plot.subtitle = element_blank(),
      panel.grid = element_blank(),
      legend.position = c(0.85, 0.2),
      legend.title = element_blank()))
gtb1plot$layers[[5]]$aes_params$alpha <- 0.4
gtb1plot$layers[[5]]$aes_params$size  <- 3
gtb1plot <- gtb1plot +
  annotate(
    "text",
    x = Inf,
    y = -Inf,
    label = "Scaling = 1, Centering = 2, SVP = 2",
    hjust = 1.1,
    vjust = -0.8,
    fontface = "bold",
    size = 3
  )
```

#### Arrange

```{r,fig.width=12, fig.height=12, message= FALSE, warning= FALSE}
gt_desired_panel <- ggarrange(gtb1plot, desiredgainplot,
  ncol = 2,
  common.legend = F, legend = "bottom",
  labels = c("A","B"),
  font.label = list(size = 10, face = "bold"), 
  align = "hv",
  hjust = -0.5,
  vjust = 1.5
)
gt_desired_panel
```

```{r, eval=FALSE}
ggsave("gt_desired_panel_LB.pdf", plot = gt_desired_panel, dpi = 600)
```

## Selection shift plot

```{r, message= FALSE, warning= FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggridges)
trait_names <- c(
  GY_BLUP   = "GY (kg/ha)",
  OHSW_BLUP = "OHSW (g)",
  PN_BLUP  = "PN (Number)",
  PH_BLUP  = "PH (cm)",
  NDF_BLUP   = "NDF (days)",
  NDM_BLUP   = "NDM (days)")
df_long <- bind_rows(
  lapply(names(BLUPgen), function(tr) {
  
    BLUPgen[[tr]] %>%
      dplyr::mutate(
        Trait  = trait_names[[tr]],
        Status = ifelse(
          GEN %in% selected$GEN,
          "Selected",
          "Not Selected")) %>%
      dplyr::select(
        GEN,
        Trait,
        Status,
        Predicted)
  }))
```

```{r, fig.width=8, fig.height=8, message= FALSE, warning= FALSE}
library(ggplot2)
library(dplyr)
library()
df_long$Trait <- factor(
  df_long$Trait,
  levels = c(
    "GY (kg/ha)",
    "OHSW (g)",
    "PN (Number)",
    "PH (cm)",
    "NDF (days)",
    "NDM (days)"
  )
)
# Cullis heritability table
h2_df <- data.frame(
  Trait = c(
    "GY (kg/ha)",
    "OHSW (g)",
    "PN (Number)",
    "PH (cm)",
    "NDF (days)",
    "NDM (days)"
  ),
  H2 = c(
    0.303,
    0.754,
    0.281,
    0.528,
    0.752,
    0.403))

h2_df$Trait <- factor(
  h2_df$Trait,
  levels = levels(df_long$Trait))
# Plot
SelectionShiftPlot <- ggplot(
  df_long,
  aes(
    x = Status,
    y = Predicted,
    fill = Status)) +
  # Violin
  geom_violin(
    alpha = 0.25,
    color = "black",
    linewidth = 1.2,
    trim = FALSE) +
  # Boxplot
  geom_boxplot(
    width = 0.12,
    alpha = 0.3,
    linewidth = 0.6,
    outlier.shape = NA) +
  # Individual genotypes
  geom_jitter(
    width = 0.08,
    size = 1.5,
    alpha = 0.45,
    color = "black") +
  # Mean
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 23,
    size = 3,
    fill = "white",
    color = "black",
    stroke = 1) +
  facet_wrap(
    ~Trait,
    scales = "free_y",
    ncol = 2, nrow = 3) +
  # Heritability annotation
  geom_text(
    data = h2_df,
    aes(
      x = 1.5,
      y = Inf,
      label = paste0(
        "H[Cullis]^2==",
        sprintf("%.3f", H2))),
    inherit.aes = FALSE,
    parse = TRUE,
    fontface = "bold",
    size = 4,
    vjust = 1.5) +
  scale_fill_manual(
    values = c(
      "Not Selected" = "orange",
      "Selected" = "purple")) +
  labs(
    x = NULL,
    y = "Predicted",
    fill = NULL) +
  ggdist::theme_ggdist(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    strip.background = element_rect(
      fill = "grey95",
      colour = "black",
      linewidth = 0.5),
    strip.text = element_text(
      face = "bold",
      size = 12),
    axis.title.y = element_text(
      face = "bold",
      size = 13),
    axis.text.x = element_text(
      face = "bold",
      size = 11),
    axis.text.y = element_text(
      face = "bold",
      size = 11),
    legend.position = "none",
    legend.text = element_text(
      face = "bold",
      size = 11),
    panel.spacing = unit(1.2, "lines"))
SelectionShiftPlot
```

```{r, eval=FALSE}
ggsave("SelectionShiftPlot.pdf", plot = SelectionShiftPlot, 
       width = 8,
       height = 8,
       dpi = 600)
```

### Density plot

```{r, eval=FALSE, echo=FALSE, message= FALSE, warning= FALSE}
BLUPgen=readxl::read_xlsx("BLUPgen_LB.xlsx",)
str(BLUPgen)
all_sheets <- readxl::excel_sheets("BLUPgen_LB.xlsx")
BLUPgen <- purrr::map(purrr::set_names(all_sheets),
                   readxl::read_excel,
                   path = "BLUPgen_LB.xlsx")


DesireIndex <- readxl::read_xlsx("Desireindex.xlsx")

selected <- DesireIndex[1:5, ]
```

```{r, eval=FALSE, echo=FALSE}
Cullis_H2 <- c(
  GY   = 0.303,
  OHSW = 0.754,
  NDF  = 0.752,
  NDM  = 0.403,
  PH   = 0.528,
  PN   = 0.281
)
plot_density_SG <- function(data_trait,
                            selected,
                            xlab,
                            h2){

  df <- as.data.frame(data_trait)

  df$Status <- factor(
    ifelse(
      df$GEN %in% selected$GEN,
      "Selected",
      "Not Selected"
    ),
    levels = c("Not Selected","Selected")
  )

  xmax <- max(df$Predicted, na.rm = TRUE)

  ggplot(df,
         aes(x = Predicted,
             fill = Status)) +

    geom_density(
      alpha = 0.6,
      color = "black",
      linewidth = 0.4,
      adjust = 5
    ) +

  annotate(
  "text",
  x = xmax,
  y = Inf,
  label = paste0(
    "H[Cullis]^2==",
    h2
  ),
  parse = TRUE,
  hjust = 1.1,
  vjust = 1.5,
  fontface = "bold",
  size = 4.5
) +
    scale_fill_manual(
      values = c(
        "Not Selected" = "#FFE5B4",
        "Selected"     = "#BFB4FF")) +
    labs(
      x = xlab,
      y = "Density",
      fill = NULL
    ) +
    theme_bw(base_size = 12) +
    theme(
      axis.title.x = element_text(face = "bold"),
      axis.title.y = element_text(face = "bold"),
      legend.position = "bottom")
}
library(patchwork)
pd1 <- plot_density_SG(
  BLUPgen$GY_BLUP,
  selected,
  expression(bold("GY (kg ha"^{-1}*")")),
  Cullis_H2["GY"])

pd2 <- plot_density_SG(
  BLUPgen$OHSW_BLUP,
  selected,
  expression(bold("OHSW (g)")),
  Cullis_H2["OHSW"])

pd3 <- plot_density_SG(
  BLUPgen$NDF_BLUP,
  selected,
  expression(bold("NDF (days)")),
  Cullis_H2["NDF"])

pd4 <- plot_density_SG(
  BLUPgen$NDM_BLUP,
  selected,
  expression(bold("NDM (days)")),
  Cullis_H2["NDM"])

pd5 <- plot_density_SG(
  BLUPgen$PH_BLUP,
  selected,
  expression(bold("PH (cm)")),
  Cullis_H2["PH"])

pd6 <- plot_density_SG(
  BLUPgen$PN_BLUP,
  selected,
  expression(bold("PN (Number)")),
  Cullis_H2["PN"])
densplot <-
  (pd1 + pd2 + pd3) /
  (pd4 + pd5 + pd6) +
  plot_annotation(
    tag_levels = "A",
    theme = theme(
      plot.tag = element_text(
        face = "bold",
        size = 14))) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r, eval=FALSE}
ggsave(plot = densplot, "densplot.pdf", width = 14, height = 8, dpi = 600)
```

## Environmental Similarity

```{r, message=FALSE, warning=FALSE}
library(EnvRtype)
library(terra)
library(raster)
library(dplyr)
library(ggplot2)
library(reshape2)

# ENVIRONMENTS
env.i <- c(
  "Piracicaba (SP)",
  "Teresina (PI)",
  "Tianguá (CE)"
)

lat <- c(-22.708333, -5.093889, -3.732222)
lon <- c(-47.636667, -42.784722, -41.012222)

plant.date <- c("2024-12-19", "2023-03-14", "2023-02-14")
harv.date  <- c("2025-04-15", "2023-07-14", "2023-06-14")

# WEATHER DATA
df.clim <- get_weather(
  env.id = env.i,
  lat = lat,
  lon = lon,
  start.day = plant.date,
  end.day = harv.date
)

# CLIMATE COVARIATES
var.clim <- c("T2M", "VPD", "PRECTOT", "GWETTOP")

df.clim$DATE <- as.Date(as.character(df.clim$YYYYMMDD), format = "%Y%m%d")
df.clim$YEAR <- as.numeric(format(df.clim$DATE, "%Y"))
df.clim$MM   <- as.numeric(format(df.clim$DATE, "%m"))
df.clim$DD   <- as.numeric(format(df.clim$DATE, "%d"))

EC <- W_matrix(
  env.data = df.clim,
  env.id = "env",
  var.id = var.clim,
  statistic = "mean"
)

EC <- data.frame(env = rownames(EC), EC, row.names = NULL)

# SOIL RASTER
soilraster <- raster(
  "C:/Users/joaop/Documents/GitHub/Kenya Maps/KenyaMaps/Soil_Raster/hwsd_domi.tif"
)

coords <- data.frame(env = env.i, lon = lon, lat = lat)

soil_values <- extract(soilraster, coords[, c("lon", "lat")])

soil_class <- data.frame(env = env.i, soil_class = soil_values)

soil_class_names <- data.frame(
  soil_class = c(1, 10),
  class_name = c("Lixisol", "Regosol")
)

soil_class <- left_join(soil_class, soil_class_names, by = "soil_class")

# MERGE CLIMATE + SOIL
EC_soil <- left_join(
  EC,
  soil_class[, c("env", "class_name")],
  by = "env"
)

# DUMMY VARIABLES (FIXED)
soil_dummy <- model.matrix(~ class_name - 1, data = EC_soil)
soil_dummy <- as.data.frame(soil_dummy)

# FINAL MATRIX
EC_final <- cbind(
  EC_soil %>% dplyr::select(-class_name),
  soil_dummy
)

EC_numeric <- EC_final %>% dplyr::select(where(is.numeric))
EC_numeric <- as.matrix(EC_numeric)

rownames(EC_numeric) <- EC_final$env

# IMPUTE NA
for (i in 1:ncol(EC_numeric)) {
  EC_numeric[is.na(EC_numeric[, i]), i] <- mean(EC_numeric[, i], na.rm = TRUE)
}

# ENVIRONMENTAL KERNEL
K_E <- env_kernel(env.data = EC_numeric)[[2]]

# HEATMAP DATA
K_df <- melt(K_E)

colnames(K_df) <- c("Env1", "Env2", "Similarity")

K_df <- K_df %>%
  dplyr::mutate(
    Env1 = factor(Env1, levels = rownames(K_E)),
    Env2 = factor(Env2, levels = colnames(K_E))
  ) %>%
  filter(as.numeric(Env1) >= as.numeric(Env2))

# PLOT
envsim <- ggplot(K_df, aes(Env1, Env2, fill = Similarity)) +

  geom_tile(color = "white", linewidth = 0.5) +

  scale_fill_gradientn(
    colors = c("purple", "#b2abd2", "#f7f7f7", "#fdb863", "orange"),
    name = "Environmental Similarity"
  ) +

  coord_equal() +

  theme_minimal(base_size = 16) +

  theme(
    panel.grid = element_blank(),

    axis.text.x = element_text(face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 12),

    axis.title = element_blank(),

    plot.title = element_text(
      face = "bold",
      hjust = 0.5,
      size = 18
    ),

    legend.position = "bottom",

    legend.title = element_text(face = "bold", size = 12),
    legend.text = element_text(face = "bold", size = 10)
  )
```

```{r, message= FALSE, warning= FALSE}
library(igraph)
library(ggraph)
library(ggplot2)
env_df <- data.frame(
  env = c(
    "Piracicaba",
    "Teresina",
    "Tianguá"
  ),
  x = c(0, 1, 0.5),
  y = c(0, 0, 0.9)
)

envsim2 = ggplot(env_df, aes(x, y)) +

  # triangle
  geom_polygon(
    fill = "#F7F7F7",
    color = "grey40",
    linewidth = 1.2,
    alpha = 0.5
  ) +

  # environment points
  geom_point(
    aes(fill = env),
    shape = 21,
    size = 20,
    color = "black",
    stroke = 0
  ) +

  # environment labels
  geom_text(
    aes(label = env),
    fontface = "bold",
    size = 5,
    nudge_y = 0,
    nudge_x = 0
  ) +

  # similarity values
  annotate(
    "text",
    x = 0.5,
    y = 0.05,
    label = round(K_E[1,2], 2),
    size = 5,
    fontface = "bold"
  ) +

  annotate(
    "text",
    x = 0.25,
    y = 0.45,
    label = round(K_E[1,3], 2),
    size = 5,
    fontface = "bold"
  ) +

  annotate(
    "text",
    x = 0.75,
    y = 0.45,
    label = round(K_E[2,3], 2),
    size = 5,
    fontface = "bold") +

  scale_fill_manual(
    values = c(
      "Piracicaba" = "purple",
      "Teresina"   = "orange",
      "Tianguá"    = "lightpink")) +

  coord_equal() +
  theme_void() +
  theme(
    legend.position = "none"
  )
envsim2
```

```{r, eval=FALSE}
ggsave("envsim2.pdf", plot = envsim2, 
       width = 8,
       height = 8,
       dpi = 600)
```
