---
title: "Leveraging G×E interaction to optimize multi-trait selection in Lima Bean"
author:
  - Gérson N.C. Ferreira$^1$†*
  - João P.S. Pavan$^1$†
  - Mauricio S. Araújo$^1$
  - Dayana R. Sousa$^1$
  - Michelle S. Nascimento$^2$
  - Vanessa G. Moura$^2$
  - José T. B. Chagas$^1$
  - Yasmim I. Retore$^1$
  - Josieli L. Silva$^2$
  - Maria S. S. Silva$^2$
  - Regina L.F. Gomes$^2$
  - Ângela C.A. Lopes$^2$
  - Maria I. Zucchi$^{3,4}$
  - José B. Pinheiro$^1$*
date: "2026-05-05"
output:
  pdf_document:
    keep_tex: true
    fig_caption: true
    number_sections: true
geometry: "left=2.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm"
fontsize: 12pt
linestretch: 1.5
editor_options:
  chunk_output_type: console
---


**Affiliations:**

**$^1$** Genetics Diversity and Breeding Laboratory, Department of Genetics, "Luiz de Queiroz" College of Agriculture (ESALQ/USP), Piracicaba, São Paulo, Brazil

**$^2$** Department of Plant Science, Federal University of Piauí, Teresina, Piauí, Brazil

**$^3$** Conservation Genetics and Genomics Group, Department of Genetics, "Luiz de Queiroz" College of Agriculture (ESALQ), University of São Paulo (USP), Piracicaba, São Paulo, Brazil

**$^4$** Secretariat of Agriculture and Food Supply of São Paulo State, Piracicaba, São Paulo, Brazil

\vspace{0.5cm}

 *Corresponding authors* * gerson.nascimento@live.com (G.N.C. Ferreira) and jbaldin@usp.br (J.B. Pinheiro)

† *These authors contributed equally to this work*

\newpage


```{r}
load("../Data/ebvMat.rda")


```



# Desire index

```{r}
# DESIRE INDEX FUNCTION

gainDiff <- function(ebvMat, b_ebv, dGain, nSel){ 
  
  take <- order(ebvMat %*% b_ebv, decreasing = TRUE)[1:nSel]
  
  mu_Sel <- colMeans(ebvMat[take, ])
  resp   <- mu_Sel - colMeans(ebvMat)
  resp   <- resp / sqrt(sum(resp^2))
  
  sum((resp - dGain)^2)
}
```


```{r}
ebvMat <- ebvMat[complete.cases(ebvMat), ]
#COV MATRIX -EBVS
library(AlphaSimR)

G_ebv <- AlphaSimR::popVar(ebvMat)
G_ebv


```

```{r}
#desired gains
desiredGain <- c(
  GY   =  2,
  OHSW =  1,
  NDF  = -1,
  NDM  = -1,
  PH   =  0,
  PN   =  1
)
desiredGain <- desiredGain / sqrt(sum(desiredGain^2))
sum(desiredGain^2)
```

```{r}
b_ebv <- solve(G_ebv) %*% desiredGain
b_ebv
```

```{r}
nSel <- 5
opt <- optim(
  par     = b_ebv,
  fn      = gainDiff,
  ebvMat  = ebvMat,
  dGain   = desiredGain,
  nSel    = nSel
)
opt_weights <- opt$par
opt_weights
```

```{r}
DesireIndex <- ebvMat %*% opt_weights

DesireIndex <- data.frame(
  GEN = rownames(ebvMat),
  DesireIndex = as.numeric(DesireIndex)
) |>
  dplyr::arrange(desc(DesireIndex))

head(DesireIndex, 10)

```

```{r}
selected <- DesireIndex[1:5, ]
selected
```

